---
title: "IBM 6540 — Midterm Exam"
subtitle: "Applied Machine Learning for Marketing"
author: "Brian Vuong"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: 1
    toc-location: right-body
    toc-title: "Questions"
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    highlight-style: github
    df-print: kable
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

# Submission Instructions {.unnumbered}

::: {.callout-important title="Read Before You Begin"}
1. **Replace "Your Name Here"** in the YAML `author` field with your full name before submitting.
2. **Render this file to HTML** — go to File → Render (or click the Render button in Positron/RStudio). Your submitted file must be the rendered `.html` file, not the `.qmd` source.
3. **All code chunks must run without errors.** Set `eval: true` for every chunk that contains your answer code. Chunks left as `eval: false` will receive zero credit for the coding portion.
4. **Written answers go in the designated prose sections** immediately below each question. Do not put written answers inside code chunks.
5. **This is an open-book, individual exam.** You may use your codebooks, notes, and textbook. You may not collaborate with other students.
6. **Submit by the deadline** as a single `.html` file via Canvas.
:::

---

# Setup {.unnumbered}

Run this chunk first. It loads all packages and recreates the datasets used throughout the exam. **Do not modify this chunk.**

```{r}
library(tidymodels)
library(tidyverse)
library(broom)
tidymodels_prefer()

# ── Customer churn dataset (from Module 1) ────────────────────────────────
# NOTE: This dataset was recalibrated so that (a) monthly_spend is now a
# function of income, tenure, and num_products (plus noise) rather than
# pure noise, so Q2's regression finds real, significant relationships,
# and (b) the churn threshold was tuned so the actual churn rate is
# genuinely close to 20%, matching Q3's narrative.
set.seed(42)
n_cust <- 1000

customer_data <- tibble(
  customer_id = 1:n_cust,
  tenure = sample(1:60, n_cust, replace = TRUE),
  income = round(rlnorm(n_cust, meanlog = 10.9, sdlog = 0.55)),
  num_products = sample(
    1:5,
    n_cust,
    replace = TRUE,
    prob = c(0.3, 0.3, 0.2, 0.1, 0.1)
  ),
  num_complaints = rpois(n_cust, lambda = 0.5),
  last_login_days = sample(1:90, n_cust, replace = TRUE),
  region = sample(
    c("West", "East", "South", "Midwest"),
    n_cust,
    replace = TRUE
  )
) |>
  mutate(
    monthly_spend = round(
      pmax(
        5,
        20 +
          0.0006 * income +
          0.30 * tenure +
          4 * num_products +
          rnorm(n_cust, 0, 30)
      ),
      2
    ),
    churn = factor(ifelse(
      0.05 *
        num_complaints +
        0.02 * last_login_days -
        0.015 * tenure -
        0.005 * monthly_spend +
        rnorm(n_cust, 0, 0.5) >
        0.84,
      "yes",
      "no"
    ))
  )

# ── Direct-mail dataset (from Modules 2 & 3) ─────────────────────────────
# NOTE: This dataset was recalibrated so that (a) the response rate is
# closer to a realistic ~20% campaign response rate, and (b) age, income,
# recency, frequency, and loyalty tier all have a meaningful relationship
# with response, so classification models built on it achieve reasonable
# (not near-chance) discrimination.
set.seed(123)
n_mail <- 3000

mail_data <- tibble(
  customer_id = paste0("C", str_pad(1:n_mail, 5, pad = "0")),
  age = round(rnorm(n_mail, mean = 45, sd = 12)),
  income = round(rlnorm(n_mail, meanlog = 10.8, sdlog = 0.6)),
  recency_days = round(rexp(n_mail, rate = 1 / 60)),
  freq_12mo = rpois(n_mail, lambda = 3),
  avg_order_amt = round(rlnorm(n_mail, meanlog = 4.2, sdlog = 0.5), 2),
  channel = sample(
    c("email", "direct_mail", "digital"),
    n_mail,
    replace = TRUE,
    prob = c(0.5, 0.3, 0.2)
  ),
  region = sample(
    c("West", "South", "Midwest", "Northeast"),
    n_mail,
    replace = TRUE
  ),
  loyalty_tier = sample(
    c("Bronze", "Silver", "Gold", "Platinum"),
    n_mail,
    replace = TRUE,
    prob = c(0.4, 0.3, 0.2, 0.1)
  ),
  responded = rbinom(
    n_mail,
    1,
    prob = plogis(
      -2.6 +
        0.03 * (age - 45) +
        0.50 * log(income / 50000) +
        0.45 * freq_12mo -
        0.02 * recency_days +
        case_when(
          loyalty_tier == "Bronze" ~ 0,
          loyalty_tier == "Silver" ~ 0.5,
          loyalty_tier == "Gold" ~ 1.1,
          loyalty_tier == "Platinum" ~ 2.0
        )
    )
  )
) |>
  mutate(
    income = if_else(runif(n_mail) < 0.08, NA_real_, income),
    avg_order_amt = if_else(runif(n_mail) < 0.05, NA_real_, avg_order_amt),
    responded = factor(responded, levels = c(1, 0), labels = c("yes", "no"))
  )

library(tidyverse)

marketing <- read_csv(
  "marketing.csv",
  show_col_types = FALSE
)

glimpse(marketing)

```

---

# Topic 1 — Machine Learning Foundations & tidymodels

*Covers: TMWR Chapters 1, 3, and 5 (Modules 1 & 2)*

## Q1 — Three Types of Models (3 points)

A marketing team is discussing three projects:

- **Project A**: They want to draw a trend line through monthly sales data to visualize growth over time.
- **Project B**: They want to test whether a new email subject line caused a statistically significant increase in open rates compared to the control group.
- **Project C**: They want to build a system that scores each prospect and predicts which ones will respond to the next mailing.

1. Classify each project (A, B, C) into one of the three model types from TMWR Chapter 1: *descriptive*, *inferential*, or *predictive*. Briefly justify each.
2. This course focuses on one of these types. Which one, and why is it the most relevant for marketing machine learning?

*Write your answer here (2–3 sentences per part).*

**1.**

* **Project A – Descriptive model:** This project summarizes and visualizes historical sales data by drawing a trend line to show how sales have changed over time. Its goal is to describe patterns in the data rather than explain causes or predict future outcomes.

* **Project B – Inferential model:** This project tests whether the new email subject line caused a statistically significant increase in open rates compared to the control group. It uses statistical inference to determine whether an observed difference is likely due to the treatment rather than random chance.

* **Project C – Predictive model:** This project builds a model that predicts which prospects are most likely to respond to a future mailing. The goal is to make accurate predictions for new, unseen data, which is the defining characteristic of predictive modeling.

**2.**

This course focuses on **predictive modeling**. Predictive models are especially important in marketing machine learning because they help businesses forecast customer behavior, such as predicting responses, purchases, or churn, allowing marketers to make more effective, data-driven decisions.


---

## Q2 — The Formula Interface and broom (5 points)

Using the `customer_data` dataset, fit a linear regression predicting `monthly_spend` from `tenure`, `num_products`, and `region`. Then answer the questions below.

::: {.callout-note title="A predictor is missing on purpose"}
`customer_data` also contains an `income` variable, which is in fact the single strongest driver of `monthly_spend` in how this data was generated. It is **intentionally excluded** from the model formula below. Keep this in mind when you interpret the R² in part 2 — a
modest R² here doesn't necessarily mean spend is unpredictable, only that this particular model is missing an important predictor.
:::

```{r q2-fit, eval=FALSE}
# Fit the model
spend_fit <- lm(
  monthly_spend ~ tenure + num_products + region,
  data = customer_data
)

# Use tidy() to extract coefficients
tidy(spend_fit)

# Use glance() to assess model fit
glance(spend_fit)

# Use augment() to get residuals and plot them
augment(spend_fit) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4, color = "#154734") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#B08D57") +
  labs(
    title = "Residuals vs. Fitted Values",
    x = "Fitted values ($)",
    y = "Residuals ($)"
  ) +
  theme_minimal()
```

1. Which predictors are statistically significant at the 0.05 level? How do you know?
2. What does the R² value tell you about the model's explanatory power? Is this model a good fit for the data?
3. What pattern (if any) do you see in the residual plot, and what might it suggest about the model?

*Write your answer here.*

1. **Statistically significant predictors:**  
At the 0.05 significance level, **tenure** (p = 1.69 × 10⁻⁶) and **num_products** (p = 8.12 × 10⁻⁶) are statistically significant because their p-values are less than 0.05. The region variables (Midwest, South, and West) are **not** significant since their p-values are all greater than 0.05.

2. **Interpretation of R²:**  
The model has an **R² of 0.0442**, meaning it explains only about **4.4% of the variation** in monthly spending. This indicates the model has **weak explanatory power** and is not a particularly good fit for the data, suggesting that other important predictors are missing.

3. **Residual plot:**  
The residual plot appears to show the residuals scattered around the zero line without a strong systematic pattern, which supports the linearity assumption. However, the low R² suggests that while the model may satisfy basic assumptions, it does not explain much of the variability in monthly spending.

---

## Q3 — The Data Budget and Stratified Splitting (4 points)

```{r q3-split, eval=FALSE}
set.seed(2024)

# Part A: Create a stratified 75/25 split on churn
churn_split <- initial_split(customer_data, prop = 0.75, strata = churn)
churn_train <- training(churn_split)
churn_test <- testing(churn_split)

# Verify churn rates in both sets
bind_rows(
  churn_train |>
    count(churn) |>
    mutate(pct = round(n / sum(n) * 100, 1), set = "Training"),
  churn_test |>
    count(churn) |>
    mutate(pct = round(n / sum(n) * 100, 1), set = "Test")
) |>
  select(set, churn, n, pct)

# Part B: What happens without stratification?
set.seed(2024)
unstrat_split <- initial_split(customer_data, prop = 0.75)
unstrat_train <- training(unstrat_split)
unstrat_test <- testing(unstrat_split)

bind_rows(
  unstrat_train |>
    count(churn) |>
    mutate(pct = round(n / sum(n) * 100, 1), set = "Train (no strata)"),
  unstrat_test |>
    count(churn) |>
    mutate(pct = round(n / sum(n) * 100, 1), set = "Test (no strata)")
) |>
  select(set, churn, n, pct)
```

1. What is the "data budget" concept from TMWR Chapter 5, and why does it matter for machine learning?
2. Compare the churn rates in training vs. test sets with and without stratification. What do you notice?
3. The churn rate in `customer_data` is approximately 20%. Explain in 2–3 sentences why stratified splitting is especially important when a class is rare — and what could go wrong without it.

*Write your answer here.*

1. The **data budget** is the limited amount of available data that must be divided among model training, tuning, and final testing. It matters because using too little data for training can weaken the model, while using too little for testing can make performance estimates unreliable.

2. With stratification, both the training and test sets have a churn rate of **19.9%**, so the class proportions are essentially identical. Without stratification, the training set has **20.3%** churn while the test set has **18.8%**, showing that random splitting can create less representative class distributions.

3. Stratified splitting is especially important when one class is rare because it preserves approximately the same class proportions in each data set. Without it, the test or training set might contain too few rare cases, causing the model to learn the minority class poorly or producing an unreliable evaluation of its performance.

---

# Topic 2 — Foundations of Logistic Regression

*Covers: Video Quiz (Steps 1-1 through 2-1)*

## Q4 — Logarithms and Marketing Data (5 points)

The `mail_data` dataset contains an `income` variable with `meanlog = 10.8` and `sdlog = 0.6`.

```{r q4-logs, eval=FALSE}
# Part A: Compute the theoretical median and mean of income
median_income <- exp(10.8)
mean_income <- exp(10.8 + 0.6^2 / 2)
cat("Theoretical median: $", round(median_income), "\n")
cat("Theoretical mean  : $", round(mean_income), "\n")

# Part B: Verify against the actual data
mail_data |>
  summarise(
    empirical_median = median(income, na.rm = TRUE),
    empirical_mean = mean(income, na.rm = TRUE)
  )

# Part C: Demonstrate why log10 base doesn't matter after normalization
income_val <- 49021
tibble(
  base = c("natural", "base10"),
  log_value = c(log(income_val), log10(income_val)),
  ratio = log(income_val) / log10(income_val)
)
```

1. Why is the mean income higher than the median income? What property of the income distribution causes this, and why is it common in real-world marketing data?
2. Your recipe uses `step_log(income, base = 10, offset = 1)` followed by `step_normalize()`. Explain why the choice of log base (10 vs. natural log) does not affect the final model, using the ratio you computed above.
3. Why does the recipe include `offset = 1` in `step_log()`? What would happen without it?

*Write your answer here.*

1. The mean income is higher than the median because the income distribution is **right-skewed (positively skewed)**, with a small number of very high-income values pulling the mean upward. This is common in marketing data because most customers have moderate incomes while a few customers earn substantially more.

2. Changing from the natural log to log base 10 only multiplies every transformed value by a constant (about **2.30**, as shown by the ratio). After `step_normalize()`, this constant scaling is removed because the data are centered and standardized, so the final model is unaffected by the choice of log base.

3. The `offset = 1` is included so that values of **0** can be safely transformed, since `log(0)` is undefined. Without the offset, any zero values would produce `-Inf` (or errors), causing problems when fitting the model.

---

## Q5 — Odds, Log(Odds), and the Logit (5 points)

```{r q5-odds, eval=FALSE}
# Response rate in mail_data
p_respond <- mean(mail_data$responded == "yes")
cat("Response probability:", round(p_respond, 4), "\n")

# Convert to odds and log(odds)
odds_respond <- p_respond / (1 - p_respond)
logodds_respond <- log(odds_respond)
cat("Odds           :", round(odds_respond, 4), "\n")
cat("Log(odds)      :", round(logodds_respond, 4), "\n")

# Symmetry of log(odds)
probs <- c(0.05, 0.50, 0.95)
odds_vec <- probs / (1 - probs)
tibble(
  probability = probs,
  odds = round(odds_vec, 4),
  log_odds = round(log(odds_vec), 4)
)
```

1. The response rate in `mail_data` is approximately 20%. Compute the **odds** of responding and express it in plain marketing language (e.g., "for every X customers who respond, Y do not").
2. The table above shows probabilities of 0.05, 0.50, and 0.95 with their corresponding log(odds). Explain why log(odds) is symmetric around zero while raw odds are not. Why does this symmetry matter for building a linear model?
3. A logistic regression model outputs `.pred_yes = 0.20` for a specific customer. Convert this to log(odds) by hand (show the formula and calculation), and explain what it means that this log(odds) value is negative.

*Write your answer here.*

1. The response probability is about **20.1%**, so the odds of responding are **0.251**. In plain marketing language, this means there is about **1 customer who responds for every 4 customers who do not respond** (odds ≈ 1:4).

2. Raw odds are not symmetric because they range from **0 to infinity**, while log(odds) transforms them to values centered around **0**. For example, probabilities of **0.05** and **0.95** have log(odds) of **-2.94** and **2.94**, respectively. This symmetry makes the relationship more suitable for a linear model because increases and decreases are treated evenly.

3. For a predicted probability of **0.20**:

\[
\text{log(odds)} = \ln\left(\frac{p}{1-p}\right)
= \ln\left(\frac{0.20}{0.80}\right)
= \ln(0.25)
\approx -1.386
\]

The negative log(odds) means the customer is **less likely to respond than not respond**, since the probability of responding is below 50%.


---

## Q6 — Odds Ratios and Coefficient Interpretation (5 points)

After fitting a logistic regression to `mail_data`, you extract the following coefficients:

| term | estimate (log-odds) |
|---|---|
| freq_12mo | 0.28 |
| recency_days | −0.19 |
| income | 0.31 |
| loyalty_tier_Platinum | 0.45 |
| loyalty_tier_Gold | 0.12 |
| loyalty_tier_Silver | 0.03 |

```{r q6-or, eval=FALSE}
# Convert log-odds coefficients to odds ratios
coef_tbl <- tibble(
  term = c(
    "freq_12mo",
    "recency_days",
    "income",
    "loyalty_tier_Platinum",
    "loyalty_tier_Gold",
    "loyalty_tier_Silver"
  ),
  estimate = c(0.28, -0.19, 0.31, 0.45, 0.12, 0.03)
) |>
  mutate(odds_ratio = exp(estimate))

coef_tbl
```

1. Interpret the odds ratio for `freq_12mo` in a complete sentence that a marketing manager could understand (do not use statistical jargon).
2. Interpret the odds ratio for `recency_days`. Why does a negative coefficient make intuitive sense for a direct-mail response model?
3. Compare the odds ratios for the three loyalty tiers. Bronze is the reference category (coefficient = 0). Which tier is the strongest predictor of response? Express the result as "Platinum customers have ___ times the odds of responding compared to Bronze customers."

*Write your answer here.*

1. Holding the other customer characteristics constant, each one-unit increase in `freq_12mo` is associated with **1.32 times the odds of responding**, or about a **32% increase** in response odds.

2. Each additional day in `recency_days` multiplies the response odds by **0.827**, which is about a **17.3% decrease** in odds. This makes sense because customers who have not purchased or interacted recently are generally less engaged and less likely to respond to direct mail.

3. Compared with Bronze customers, Silver customers have **1.03 times**, Gold customers have **1.13 times**, and Platinum customers have **1.57 times** the odds of responding. Therefore, Platinum is the strongest predictor: **Platinum customers have 1.57 times the odds of responding compared with Bronze customers.**

---

## Q7 — Logistic vs. Linear Regression (5 points)

```{r q7-sigmoid, eval=FALSE}
# Sigmoid function at various linear predictor values
tibble(
  linear_predictor = c(-4, -2, -1, 0, 1, 2, 4)
) |>
  mutate(
    probability = plogis(linear_predictor)
  )
```

1. Why can't we use linear regression to model a binary outcome like `responded` (yes/no)? Give one specific example of what goes wrong, using numbers from the `mail_data` response rate (≈ 20%).
2. The sigmoid function is defined as $P = \frac{1}{1+e^{-z}}$ where $z$ is the linear predictor. When $z = 0$, what probability does `plogis(0)` return, and why is that mathematically necessary?
3. Logistic regression uses **maximum likelihood** rather than **least squares** to fit the model. In plain English, what is maximum likelihood estimating — what is it trying to maximize and why?

*Write your answer here.*

1. Linear regression is inappropriate for a binary outcome because it can produce predicted probabilities below 0 or above 1. For example, starting from the roughly **20%** response rate, a linear model could predict a probability of **-0.05** for a low-scoring customer or **1.10** for a high-scoring customer, neither of which is possible; logistic regression keeps predictions between 0 and 1.

2. When \(z = 0\), `plogis(0)` returns **0.50** because

\[
p = \frac{1}{1 + e^{-0}} = \frac{1}{2} = 0.5.
\]

This is mathematically necessary because zero log-odds means the odds are \(1:1\), so responding and not responding are equally likely.

3. Maximum likelihood chooses the coefficients that make the outcomes actually observed in the data as probable as possible. In this case, it tries to assign high predicted probabilities to customers who responded and low predicted probabilities to customers who did not respond, selecting the coefficient values that maximize the combined likelihood of all those observed yes/no results.


---

# Topic 3 — Data Splitting & Preprocessing

*Covers: TMWR Chapters 5 and 8 (Module 2 Codebook)*

## Q8 — Recipe Design and Step Order (6 points)

Below is a recipe for the direct-mail dataset. Answer the questions based on this recipe.

```{r q8-recipe, eval=FALSE}
set.seed(617)
mail_split <- initial_split(mail_data, prop = 0.80, strata = responded)
mail_train <- training(mail_split)
mail_test <- testing(mail_split)

mail_rec <- recipe(
  responded ~ customer_id +
    age +
    income +
    recency_days +
    freq_12mo +
    avg_order_amt +
    channel +
    region +
    loyalty_tier,
  data = mail_train
) |>
  update_role(customer_id, new_role = "ID") |>
  step_impute_median(all_numeric_predictors()) |> # Step 1
  step_log(
    income,
    avg_order_amt,
    recency_days, # Step 2
    base = 10,
    offset = 1
  ) |>
  step_normalize(all_numeric_predictors()) |> # Step 3
  step_dummy(all_nominal_predictors()) |> # Step 4
  step_zv(all_predictors()) # Step 5

mail_prep <- prep(mail_rec)
tidy(mail_prep)
tidy(mail_prep, number = 1) # imputation values
tidy(mail_prep, number = 3) # normalization statistics
```

1. Why must `step_impute_median()` come **before** `step_log()`? What would happen to the output if the order were reversed?
2. The output of `tidy(mail_prep, number = 3)` shows a mean close to 4.69 for `income`. Explain why the mean is 4.69 and not 10.8 (the `meanlog` used to generate income). Show the calculation.
3. What is the purpose of including `customer_id` in the formula and then calling `update_role(customer_id, new_role = "ID")`? What would happen if `update_role()` were removed but `customer_id` stayed in the formula?

*Write your answer here.*

1. `step_impute_median()` should come before `step_log()` so missing numeric values are filled on their original scale before any transformation is applied. If the order were reversed, missing values would remain `NA` during the log step and would later be imputed using medians calculated on the logged scale, which could change the imputed and normalized output and may cause problems for preprocessing steps that require complete values.

2. The value `meanlog = 10.8` refers to the mean of the **natural logarithm** of income, but the recipe converts income using a base-10 logarithm. Using the change-of-base formula,

\[
\log_{10}(x)=\frac{\ln(x)}{\ln(10)}
\]

so the expected transformed mean is

\[
\frac{10.8}{\ln(10)}
=\frac{10.8}{2.3026}
\approx 4.69.
\]

The `offset = 1` has almost no effect here because the income values are large.

3. Including `customer_id` in the formula keeps it available for identifying and matching rows, while `update_role(customer_id, new_role = "ID")` prevents it from being used as a predictor. Without `update_role()`, the customer ID would be treated as a model input, potentially causing meaningless patterns, overfitting, or a very large number of dummy variables if the ID is categorical.



---

## Q9 — prep(), bake(), and Data Leakage (6 points)

```{r q9-leakage, eval=FALSE}
# Correct approach: prep on training, bake on test
prep_correct <- prep(mail_rec)
test_correct <- bake(prep_correct, new_data = mail_test)

# Incorrect approach: prep on test set (data leakage)
prep_leaked <- prep(mail_rec, training = mail_test)
test_leaked <- bake(prep_leaked, new_data = mail_test)

# Compare the normalization mean for income between the two approaches
bind_rows(
  tidy(prep_correct, number = 3) |>
    filter(terms == "income", statistic == "mean") |>
    mutate(approach = "correct (prep on train)"),
  tidy(prep_leaked, number = 3) |>
    filter(terms == "income", statistic == "mean") |>
    mutate(approach = "leaked (prep on test)")
) |>
  select(approach, terms, statistic, value)
```

1. Explain what **data leakage** means in the context of `prep()` and `bake()`. Why is prepping on the test set a form of leakage?
2. The two approaches produce slightly different normalization statistics for `income`. Why are they different, and why does even a small difference matter for model evaluation?
3. In practice, the performance gap caused by this leakage is often small. Does that mean it is acceptable? Explain your reasoning.

*Write your answer here.*

1. **Data leakage** occurs when information from the test set influences how the model or preprocessing steps are trained. `prep()` estimates values such as medians, means, standard deviations, and factor levels, so running it on `mail_test` allows the preprocessing pipeline to learn from data that should remain completely unseen until final evaluation.

2. The normalization means differ because the training and test sets contain slightly different customers and therefore have different income distributions; here, the training-based mean is **4.69**, while the test-based mean is **4.68**. Even a small difference matters because using the test statistic contaminates the evaluation and makes the reported performance less representative of how the model will perform on truly new data.

3. No, the leakage is not acceptable even when the measured performance difference is small. The test set is meant to provide an independent estimate of future performance, and using any information from it breaks that independence and can produce biased or overly favorable results.


---

## Q10 — Workflow and Regularization (6 points)

```{r q10-workflow, eval=FALSE}
# Fit a Lasso logistic regression workflow
lr_spec <- logistic_reg(penalty = 0.01, mixture = 1) |>
  set_engine("glmnet") |>
  set_mode("classification")

mail_wf <- workflow() |> add_recipe(mail_rec) |> add_model(lr_spec)
mail_fit <- fit(mail_wf, data = mail_train)

# Extract coefficients
mail_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(term != "(Intercept)") |>
  arrange(desc(abs(estimate)))
```

1. What does `mixture = 1` specify in the model? What regularization method does this correspond to, and what is its distinguishing property compared to Ridge regression?
2. Several predictors have `estimate = 0` in the output (for example, `region` and `avg_order_amt`). What did the Lasso do to these predictors, and why is this useful in a marketing context with many potential predictors?
3. Why is a `workflow()` preferred over calling `prep()`, `bake()`, and `fit()` separately? Give one concrete reason related to preventing mistakes in a real modeling project.

*Write your answer here.*

1. `mixture = 1` specifies **pure Lasso regularization**, which uses an L1 penalty. Unlike Ridge regression, which usually only shrinks coefficients toward zero, Lasso can reduce some coefficients to **exactly zero** and effectively remove those predictors from the model.

2. The Lasso set the coefficients for predictors such as `avg_order_amt` and several `region` variables to zero, meaning they were not retained as useful predictors at the chosen penalty level. This automatic feature selection is helpful in marketing data with many potential predictors because it produces a simpler model and can reduce overfitting.

3. A `workflow()` keeps the preprocessing recipe and model together so the same steps are applied consistently during training, testing, resampling, and future prediction. For example, it prevents accidentally calculating normalization statistics on the test set or applying different dummy-variable encoding to the training and test data.


---

# Topic 4 — Model Evaluation & Resampling

*Covers: TMWR Chapters 9 and 10 (Module 3 Codebook)*

## Q11 — Regression Metrics (9 points)

Unlike the other datasets in this exam, `marketing` is **real, not simulated** — daily performance data for 11 marketing campaigns over 28 days, from the ["Digital Marketing Metrics & KPIs to Measure (SQL)"](https://www.kaggle.com/datasets/0d4c0018901ff857a480079fdcf4f9e7c5ddce3367ecd3c05dd757c3a1c2c12d)
on Kaggle. Using this dataset, fit a linear regression predicting daily `revenue` from `mark_spent`, `clicks`, and `leads`. Then answer the questions below.

::: {.callout-note title="A quirk of real data"}
Because `lm()` has no constraint keeping predictions non-negative, some campaign-days in the test set will get a **negative predicted revenue** — obviously impossible, since revenue can't be negative. This is the same fundamental issue as Q7's discussion of linear regression for
binary outcomes: an unconstrained linear model can output values outside the range the outcome can actually take.
:::

```{r q11-reg, eval=FALSE}
set.seed(999)
marketing_split <- initial_split(marketing, prop = 0.80)
marketing_train <- training(marketing_split)
marketing_test <- testing(marketing_split)

reg_rec <- recipe(
  Revenue ~ Clicks + Leads,
  data = marketing_train
) |>
  step_normalize(all_numeric_predictors())

reg_wf <- workflow() |>
  add_recipe(reg_rec) |>
  add_model(linear_reg() |> set_engine("lm"))

reg_fit <- fit(reg_wf, data = marketing_train)
reg_preds <- augment(reg_fit, new_data = marketing_test)

# Compute metrics
reg_preds |>
  metric_set(rmse, rsq, mae)(truth = Revenue, estimate = .pred)

# Add a "whale" campaign-day and see how metrics change
whale <- tibble(Revenue = 2800000, .pred = 200000)
reg_preds_whale <- bind_rows(
  reg_preds |> select(Revenue, .pred),
  whale
)
reg_preds_whale |>
  metric_set(rmse, mae)(truth = Revenue, estimate = .pred)
```

1. Explain in plain English what **RMSE** measures. How is it different from **MAE**, and why does RMSE penalize large errors more?
2. After adding the "whale" campaign-day (actual revenue = \$2,800,000, predicted = \$200,000), which metric — RMSE or MAE — changes more in percentage terms? Why is this the expected result?
3. If you were reporting model performance to a VP of Marketing who is trying to plan next month's marketing budget, which metric (RMSE, MAE, or R²) would you report, and why? Answer in 2–3 sentences.

*Write your answer here.*

1. **RMSE** measures the typical size of the model’s prediction errors in the same units as revenue, but it gives extra weight to large mistakes. **MAE** is the average absolute difference between actual and predicted revenue, treating each error proportionally; RMSE penalizes large errors more because the errors are squared before being averaged.

2. RMSE increases from **$288,413 to $289,453**, an increase of about **0.36%**:

\[
\frac{289453-288413}{288413}\times 100 \approx 0.36\%
\]

MAE increases from **$188,322 to $188,539**, an increase of about **0.12%**. Therefore, **RMSE changes more** because the whale campaign’s very large prediction error is squared, giving it greater influence than it has on MAE.

3. I would report **MAE** because it tells the VP that the model’s predictions are off by about **$188,322 on average**, which is directly interpretable for budgeting. MAE is also less distorted by an unusually large campaign than RMSE, while \(R^2\) describes explained variation rather than the expected error in dollars.



---

## Q12 — Confusion Matrix and Classification Metrics (11 points)

```{r q12-conf, eval=FALSE}
# Fit classification model and generate predictions
set.seed(617)
mail_split2 <- initial_split(mail_data, prop = 0.80, strata = responded)
mail_train2 <- training(mail_split2)
mail_test2 <- testing(mail_split2)

lr_wf2 <- workflow() |>
  add_recipe(
    recipe(
      responded ~ customer_id +
        age +
        income +
        recency_days +
        freq_12mo +
        avg_order_amt +
        channel +
        region +
        loyalty_tier,
      data = mail_train2
    ) |>
      update_role(customer_id, new_role = "ID") |>
      step_impute_median(all_numeric_predictors()) |>
      step_log(income, avg_order_amt, recency_days, base = 10, offset = 1) |>
      step_normalize(all_numeric_predictors()) |>
      step_dummy(all_nominal_predictors()) |>
      step_zv(all_predictors())
  ) |>
  add_model(
    logistic_reg(penalty = 0.01, mixture = 1) |>
      set_engine("glmnet") |>
      set_mode("classification")
  )

lr_fit2 <- fit(lr_wf2, data = mail_train2)
test_preds2 <- augment(lr_fit2, new_data = mail_test2)

# Confusion matrix
conf_mat(test_preds2, truth = responded, estimate = .pred_class)

# Metrics
test_preds2 |>
  metric_set(accuracy, sens, spec, precision)(
    truth = responded,
    estimate = .pred_class
  )

# Lower threshold
preds_low_threshold <- test_preds2 |>
  mutate(
    .pred_class_03 = factor(
      if_else(.pred_yes > 0.30, "yes", "no"),
      levels = c("yes", "no")
    )
  )

preds_low_threshold |>
  metric_set(sens, precision)(
    truth = responded,
    estimate = .pred_class_03
  )
```

1. From the confusion matrix, identify the **false negatives** and explain what they represent in the context of a direct-mail campaign. What is the business cost of a false negative?
2. The model achieves roughly 82% accuracy at the default 0.5 threshold. Given that a "dumb" model that always predicts "no" would score close to 80% accuracy (since the response rate is only about 20%), how much real lift is the model providing? Why is accuracy alone a limited way to describe that lift?
3. When the threshold is lowered from 0.5 to 0.3, sensitivity increases and precision decreases. Explain this tradeoff in marketing terms: what does the campaign manager gain and what do they give up?

*Write your answer here.*


1. The **false negatives are 94 customers**: the model predicted “no,” but they actually responded “yes.” In a direct-mail campaign, these are customers the company may choose not to contact even though they would have responded, creating an opportunity cost from lost sales or revenue.

2. The model’s accuracy of **82.4%** is only about **2.4 percentage points higher** than an always-“no” model with roughly 80% accuracy. Accuracy alone hides the limited lift because the classes are imbalanced: the model’s sensitivity is only **22.3%**, meaning it identifies fewer than one-quarter of the actual responders and misses most of them.

3. Lowering the threshold from 0.5 to 0.3 raises sensitivity from **22.3% to 48.8%**, so the campaign manager identifies more potential responders and reduces false negatives. However, precision falls from **69.2% to 47.2%**, meaning more nonresponders will also receive mail, increasing campaign costs and reducing the response rate among targeted customers.


---

## Q13 — ROC Curve and AUC (10 points)

```{r q13-roc, eval=FALSE}
# ROC curve
roc_data <- test_preds2 |>
  roc_curve(truth = responded, .pred_yes)

autoplot(roc_data) +
  labs(title = "ROC Curve — Direct-Mail Response Model") +
  theme_minimal()

# AUC value
auc_val <- test_preds2 |>
  roc_auc(truth = responded, .pred_yes)
auc_val
```

1. What does the ROC curve show that a single confusion matrix does not? Why is this a more complete view of model performance?
2. Complete this sentence using the AUC value you computed: "If we randomly pick one customer who **would** respond and one who **would not**, this model assigns a higher predicted probability to the true responder \_\_\_% of the time." Explain why AUC = 0.5 would be equivalent to random guessing.
3. A colleague proposes targeting only the customers in the **top 20%** of predicted probabilities (the highest `.pred_yes` scores). How would you use the ROC curve to estimate what sensitivity you would achieve at that targeting rate? Describe the process (no code required).

*Write your answer here.*

1. A single confusion matrix evaluates the model at only one classification threshold, while the ROC curve shows the tradeoff between sensitivity and false-positive rate across all possible thresholds. This gives a more complete view because it shows how model performance changes as the campaign becomes more or less aggressive in targeting customers.

2. If we randomly select one customer who would respond and one who would not, the model assigns a higher predicted probability to the true responder **78.6% of the time**. An AUC of 0.5 represents random guessing because the model would rank the responder above the nonresponder only half the time.

3. First, rank customers from highest to lowest predicted response probability and find the cutoff that includes the top **20%** of customers. Then locate the point on the ROC curve corresponding to that probability cutoff and read its sensitivity from the vertical axis; this is the estimated percentage of all actual responders captured by the campaign. The targeting rate should not be read directly as \(x=0.20\), because the ROC curve’s horizontal axis is the false-positive rate, not the percentage of all customers targeted.



---

## Q14 — Resubstitution and Cross-Validation (10 points)

```{r q14-resub, eval=FALSE}
# Compare resubstitution vs. held-out accuracy
train_preds2 <- augment(lr_fit2, new_data = mail_train2)

bind_rows(
  train_preds2 |>
    roc_auc(truth = responded, .pred_yes) |>
    mutate(data = "training (resubstitution)"),
  test_preds2 |>
    roc_auc(truth = responded, .pred_yes) |>
    mutate(data = "test (held out)")
) |>
  select(data, .estimate)

# 10-fold cross-validation
set.seed(2024)
mail_folds <- vfold_cv(mail_train2, v = 10, strata = responded)

cv_results <- fit_resamples(
  lr_wf2,
  resamples = mail_folds,
  metrics = metric_set(roc_auc, accuracy, sens),
  control = control_resamples(save_pred = TRUE)
)

collect_metrics(cv_results)
```

1. Define **resubstitution** and explain why it tends to produce optimistic performance estimates. Is the gap large or small for this Lasso logistic regression, and why?
2. Describe how **10-fold cross-validation** works in 3–4 sentences. Why does it produce a more trustworthy performance estimate than a single train/test split?
3. The cross-validated `roc_auc` comes with a **standard error** (`std_err` column). What does this standard error represent, and why is it useful information that a single test-set AUC cannot provide?

*Write your answer here.*

1. **Resubstitution** evaluates a model on the same data used to fit it, so the estimate is usually optimistic because the model has already learned patterns in those observations. Here, the gap is small: training AUC is **0.796** and held-out test AUC is **0.786**, a difference of only **0.010**, likely because Lasso regularization limits model complexity and reduces overfitting.

2. In **10-fold cross-validation**, the training data are divided into 10 approximately equal folds. The model is trained on nine folds and evaluated on the remaining fold, repeating the process until every fold has served as the validation set; the 10 results are then averaged. This is more trustworthy than one train/test split because every observation is used for both training and validation, and the result depends less on one fortunate or unfortunate split.

3. The `std_err` measures the uncertainty in the mean cross-validated performance estimate caused by variation across the 10 folds. For ROC AUC, the mean is **0.791** with a standard error of about **0.0105**, indicating how precisely the average AUC has been estimated; a single held-out AUC provides only one value and does not show how much performance might vary across different samples.


---

## Q15 — Comparing Resampling Methods and Model Flexibility (10 points)

```{r q15-flex, eval=FALSE}


install.packages("ranger")
library(ranger)


# Compare 5-fold vs 10-fold CV
set.seed(2024)
folds_5 <- vfold_cv(mail_train2, v = 5, strata = responded)

cv_5fold <- fit_resamples(
  lr_wf2,
  resamples = folds_5,
  metrics = metric_set(roc_auc)
)

bind_rows(
  collect_metrics(cv_results) |>
    filter(.metric == "roc_auc") |>
    mutate(method = "10-fold CV"),
  collect_metrics(cv_5fold) |>
    filter(.metric == "roc_auc") |>
    mutate(method = "5-fold CV")
) |>
  select(method, mean, std_err)

# Resubstitution gap: Lasso vs. Random Forest, measured with AUC
# (accuracy is a poor choice here since ~80% of customers are "no" in
#  both train and test, so accuracy alone can hide overfitting)
rf_spec <- rand_forest() |>
  set_engine("ranger") |>
  set_mode("classification")

rf_wf <- lr_wf2 |> update_model(rf_spec)
rf_fit <- fit(rf_wf, data = mail_train2)

rf_train_preds <- augment(rf_fit, new_data = mail_train2)
rf_test_preds <- augment(rf_fit, new_data = mail_test2)

bind_rows(
  rf_train_preds |>
    roc_auc(truth = responded, .pred_yes) |>
    mutate(model = "Random Forest", data = "train"),
  rf_test_preds |>
    roc_auc(truth = responded, .pred_yes) |>
    mutate(model = "Random Forest", data = "test"),
  train_preds2 |>
    roc_auc(truth = responded, .pred_yes) |>
    mutate(model = "Lasso LR", data = "train"),
  test_preds2 |>
    roc_auc(truth = responded, .pred_yes) |>
    mutate(model = "Lasso LR", data = "test")
) |>
  select(model, data, .estimate) |>
  pivot_wider(names_from = data, values_from = .estimate) |>
  mutate(gap = round(train - test, 4))
```

1. When the number of folds decreases from 10 to 5, what typically happens to the **standard error** of the cross-validated metric? Explain why in terms of the number of estimates being averaged.
2. The random forest shows a much larger resubstitution AUC gap (train − test) than the Lasso logistic regression. Explain why more flexible models tend to overfit more, and what the Lasso's penalty is doing to prevent this.
3. Given what you observed about the resubstitution gap, explain why it would be **unfair** to compare a Lasso logistic regression and a random forest using their training-set performance. What should you use instead?

*Write your answer here.*

1. Decreasing from 10 folds to 5 folds would typically make the **standard error larger** because the reported mean is based on fewer performance estimates. In this output, the standard errors are nearly identical—**0.0105** for 10-fold CV and **0.0104** for 5-fold CV—so the tiny decrease is just sampling variation rather than a meaningful improvement in precision.

2. The random forest has a much larger resubstitution gap: its training AUC is **0.997**, but its test AUC is **0.797**, producing a gap of **0.200**. Random forests are flexible enough to learn complex patterns and noise in the training data, while Lasso shrinks coefficients and sets some to zero, limiting complexity and producing a much smaller gap of about **0.009**.

3. Comparing the models using training performance would be unfair because the more flexible random forest can fit the training observations almost perfectly, even when that performance does not transfer to new customers. The models should instead be compared using the **same held-out test set** or, preferably, their performance across the **same cross-validation folds**.


---

# Bonus Question (+5 points) {.unnumbered}

## BQ — End-to-End Critical Thinking

A marketing analyst at a retail company runs the following code and reports:


"Our model achieves an AUC of 0.99 — nearly perfect performance! We should deploy this immediately to score all prospects for the next mailing."

Identify **at least three distinct problems** with this analyst's approach and the conclusion they drew. For each problem, explain what went wrong and what the correct approach would be.

*Write your answer here (3–5 sentences per problem).*

1. **The model is evaluated on the same data used to train it.** The AUC of **0.996** is a resubstitution estimate, so it measures how well the random forest ranks observations it has already seen rather than new prospects. This is highly optimistic, especially because the earlier results showed that the random forest’s AUC fell from about **0.997 on training data to 0.797 on test data**. The correct approach is to estimate performance using cross-validation or an untouched test set before fitting the final model to the full dataset.

2. **The analyst ignores overfitting and model flexibility.** Random forests can learn complicated relationships and noise in the training data, allowing nearly perfect training performance without equally strong performance on unseen customers. The large gap between training and test AUC is evidence that this model has overfit. Model complexity and hyperparameters should be selected through resampling, and the final reported performance should come from held-out data.

3. **AUC alone is not enough to justify deployment.** AUC evaluates ranking performance across all possible classification thresholds, but it does not specify which customers should receive the mailing or account for mailing costs, expected profit, false positives, and false negatives. Before deployment, the analyst should select a threshold or targeting percentage using business costs and then report relevant measures such as sensitivity, precision, expected response rate, and campaign profit.

4. **Immediate full deployment is premature.** Even a model with good held-out performance may perform differently on future prospects because customer behavior, campaign design, and market conditions can change. The company should first conduct a limited pilot or randomized campaign to verify that model-based targeting creates incremental value compared with the current strategy. Performance and data drift should then be monitored after deployment.

---