---
title: "W10 - Machine Learning Deployment"
subtitle: "Predicting Product Returns with a Boosted Tree Classifier"
author: "Brian Vuong"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: true
    embed-resources: true
    theme: cosmo
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  freeze: false
---

# 1. Project & Data Overview

## 1.1 Project Description

This dashboard deploys the boosted tree model developed for **RO3** of our group's Shopify sales project: predicting whether a purchased item will be returned (`is_returned`). Discount percentage and shipping cost are framed as **pre-purchase levers** a retailer can adjust before an order ships; customer rating is framed as a **post-purchase monitoring signal**. Product category sets the baseline risk context.

The calculator below lets you enter a hypothetical order's characteristics and see the model's predicted return probability instantly, entirely in your browser — no server, no API calls.

## 1.2 Data Description

- **Source:** `shopify_sales_dataset_ml_eda.csv`, 60,000 Shopify orders
- **Outcome:** `is_returned` (binary; \~15% positive class)
- **Predictors used by this model:**
  - `product_category` — categorical, 7 levels
  - `discount_percent` — numeric, 0-100
  - `shipping_cost` — numeric, dollars
  - `rating` — numeric, 1.0-5.0

# 2. Model Deployment

## 2.1 R Code — Training & Export

```{r}
#| label: setup

library(tidyverse)
library(tidymodels)
library(xgboost)
library(jsonlite)

tidymodels_prefer()

df <- read_csv("shopify_sales_dataset_ml_eda.csv") %>%
  mutate(
    is_returned      = factor(is_returned, levels = c(0, 1), labels = c("No", "Yes")),
    product_category = as.factor(product_category)
  )
```

```{r}
#| label: fit-model

set.seed(123)
data_split <- initial_split(df, prop = 0.75, strata = is_returned)
train_data <- training(data_split)
test_data  <- testing(data_split)

# No step_normalize() -- unnecessary for tree-based models, cut for speed
returns_recipe <- recipe(
    is_returned ~ product_category + discount_percent + shipping_cost + rating,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

# Fast-render spec: fixed trees = 50, no tuning
boost_spec <- boost_tree(trees = 50) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

boost_wf  <- workflow() %>% add_recipe(returns_recipe) %>% add_model(boost_spec)
boost_fit <- fit(boost_wf, data = train_data)

# Quick holdout check -- single split, no resampling, kept lightweight
boost_preds <- augment(boost_fit, new_data = test_data)
roc_auc(boost_preds, truth = is_returned, .pred_Yes, event_level = "second")
```

```{r}
#| label: export-for-js

booster <- extract_fit_engine(boost_fit)
tree_dump <- xgb.dump(booster, dump_format = "json")

baked_cols <- returns_recipe %>%
  prep() %>%
  bake(new_data = NULL) %>%
  select(-is_returned) %>%
  colnames()

category_levels <- levels(df$product_category)

# Build one JSON blob with everything the browser calculator needs
model_payload <- list(
  trees      = lapply(tree_dump, jsonlite::fromJSON, simplifyVector = FALSE),
  features   = baked_cols,
  categories = category_levels
)
payload_json <- jsonlite::toJSON(model_payload, auto_unbox = TRUE)
```

```{r}
#| label: inject-data
#| echo: false
#| results: asis

cat(paste0('<script>\nconst modelPayload = ', payload_json, ';\n</script>'))
```


```{r}
#| eval: false

returns_recipe <- recipe(
    is_returned ~ product_category + discount_percent + shipping_cost + rating,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

boost_spec <- boost_tree(trees = 50) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

boost_wf  <- workflow() %>% add_recipe(returns_recipe) %>% add_model(boost_spec)
boost_fit <- fit(boost_wf, data = train_data)
```

## 2.2 Interactive Calculator (Live in Browser)

```{=html}
<div style="max-width:500px; font-family:sans-serif;">
  <label>Product Category</label><br>
  <select id="categoryInput" style="width:100%; margin-bottom:10px;"></select>

  <label>Discount % (<span id="discountVal">20</span>)</label><br>
  <input type="range" id="discountInput" min="0" max="100" value="20" style="width:100%; margin-bottom:10px;">

  <label>Shipping Cost $ (<span id="shippingVal">8</span>)</label><br>
  <input type="range" id="shippingInput" min="0" max="50" step="0.5" value="8" style="width:100%; margin-bottom:10px;">

  <label>Customer Rating (<span id="ratingVal">3.5</span>)</label><br>
  <input type="range" id="ratingInput" min="1" max="5" step="0.1" value="3.5" style="width:100%; margin-bottom:10px;">

  <div style="font-size:1.4em; margin-top:1em;">
    Predicted return probability: <b><span id="probOut">--</span>%</b>
  </div>
  <div style="width:100%; background:#eee; border-radius:6px; height:20px; margin-top:0.5em;">
    <div id="probBar" style="width:0%; background:#2e7d32; height:100%; border-radius:6px;"></div>
  </div>
</div>

<script>
function scoreTree(node, features) {
  if ("leaf" in node) return node.leaf;
  const val = features[node.split] ?? 0;
  const nextId = val < node.split_condition ? node.yes : node.no;
  const next = node.children.find(c => c.nodeid === nextId);
  return scoreTree(next, features);
}

function buildFeatureVector(category, discount, shipping, rating) {
  const f = {};
  for (const name of modelPayload.features) f[name] = 0;
  f["discount_percent"] = discount;
  f["shipping_cost"] = shipping;
  f["rating"] = rating;
  const dummyCol = modelPayload.features.find(n => n === `product_category_${category}`);
  if (dummyCol) f[dummyCol] = 1;
  return f;
}

function updateCalculator() {
  const category = document.getElementById("categoryInput").value;
  const discount = parseFloat(document.getElementById("discountInput").value);
  const shipping = parseFloat(document.getElementById("shippingInput").value);
  const rating = parseFloat(document.getElementById("ratingInput").value);

  document.getElementById("discountVal").textContent = discount;
  document.getElementById("shippingVal").textContent = shipping;
  document.getElementById("ratingVal").textContent = rating;

  const features = buildFeatureVector(category, discount, shipping, rating);
  const rawScore = modelPayload.trees.reduce((sum, t) => sum + scoreTree(t, features), 0);
  const prob = 1 / (1 + Math.exp(-rawScore));

  document.getElementById("probOut").textContent = (prob * 100).toFixed(1);
  const bar = document.getElementById("probBar");
  bar.style.width = (prob * 100).toFixed(1) + "%";
  bar.style.background = prob > 0.5 ? "#c0392b" : "#2e7d32";
}

document.addEventListener("DOMContentLoaded", function() {
  const select = document.getElementById("categoryInput");
  modelPayload.categories.forEach(cat => {
    const opt = document.createElement("option");
    opt.value = cat;
    opt.textContent = cat;
    select.appendChild(opt);
  });
  ["categoryInput", "discountInput", "shippingInput", "ratingInput"].forEach(id => {
    document.getElementById(id).addEventListener("input", updateCalculator);
  });
  updateCalculator();
});
</script>
```

## 2.3 Experiment Log


| Scenario | Category | Discount % | Shipping \$ | Rating | Predicted P(Return) |
|------------|------------|------------|------------|------------|------------|
| Low risk | Electronics | 5 | 5 | 4.8 | *\[fill in\]* |
| High discount, low rating | Fashion | 60 | 10 | 1.8 | *\[fill in\]* |
| High shipping cost | Footwear | 20 | 35 | 3.5 | *\[fill in\]* |
| Baseline / average order | Accessories | 20 | 8 | 3.5 | *\[fill in\]* |

**Observations:** *\[Does the gap between the low-risk and high-discount/ low-rating scenarios support the RO3 hypothesis that high discount + low rating drives returns? Does the JS calculator's output match R's predict() closely enough to trust the deployed version?\]*

## 2.4 Deployed Model on the Web

Published via Githib Pages