---
title: "W10 - Machine Learning Deployment"
subtitle: "Predicting Product Returns with a Boosted Tree Classifier"
author: "Brian Vuong"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: true
    theme: cosmo
    code-fold: true
server: shiny
execute:
  echo: true
  warning: false
  message: false
  freeze: false
---

# 1. Project & Data Overview

## 1.1 Project Description

This dashboard deploys the boosted tree model developed for **RO3** of our group's Shopify sales project: predicting whether a purchased item will be returned (`is_returned`). Discount percentage and shipping cost are framed as **pre-purchase levers** a retailer can adjust before an order ships; customer rating is framed as a **post-purchase monitoring signal**. Product category sets the baseline risk context.

The calculator below lets you enter a hypothetical order's characteristics and see the model's predicted return probability instantly using a live R Shiny server.

## 1.2 Data Description

- **Source:** `shopify_sales_dataset_ml_eda.csv`, 60,000 Shopify orders
- **Outcome:** `is_returned` (binary; ~15% positive class)
- **Predictors used by this model:**
  - `product_category` — categorical, 7 levels
  - `discount_percent` — numeric, 0-100
  - `shipping_cost` — numeric, dollars
  - `rating` — numeric, 1.0-5.0

# 2. Model Deployment

## 2.1 R Code — Training & Fit

```{r}
#| label: setup

library(tidyverse)
library(tidymodels)
library(xgboost)
library(shiny)

tidymodels_prefer()

df <- read_csv("shopify_sales_dataset_ml_eda.csv") %>%
  mutate(
    is_returned      = factor(is_returned, levels = c(0, 1), labels = c("No", "Yes")),
    product_category = as.factor(product_category)
  )
```

```{r}
#| label: fit-model

set.seed(123)
data_split <- initial_split(df, prop = 0.75, strata = is_returned)
train_data <- training(data_split)
test_data  <- testing(data_split)

# No step_normalize() -- unnecessary for tree-based models, cut for speed
returns_recipe <- recipe(
    is_returned ~ product_category + discount_percent + shipping_cost + rating,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

# Fast-render spec: fixed trees = 50, no tuning
boost_spec <- boost_tree(trees = 50) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

boost_wf  <- workflow() %>% add_recipe(returns_recipe) %>% add_model(boost_spec)
boost_fit <- fit(boost_wf, data = train_data)

# Quick holdout check -- single split, no resampling, kept lightweight
boost_preds <- augment(boost_fit, new_data = test_data)
roc_auc(boost_preds, truth = is_returned, .pred_Yes, event_level = "second")
```

## 2.2 Interactive Calculator (Shiny)

```{r}
#| label: shiny-ui

selectInput("categoryInput", "Product Category", choices = levels(df$product_category))
sliderInput("discountInput", "Discount %", min = 0, max = 100, value = 20)
sliderInput("shippingInput", "Shipping Cost ($)", min = 0, max = 50, value = 8, step = 0.5)
sliderInput("ratingInput", "Customer Rating", min = 1, max = 5, value = 3.5, step = 0.1)

h4("Predicted Return Probability:")
uiOutput("probDisplay")
```

```{r}
#| context: server

output$probDisplay <- renderUI({
  req(input$categoryInput, input$discountInput, input$shippingInput, input$ratingInput)
  
  # Format single-row dataframe matching recipe expectations
  new_order <- tibble(
    product_category = factor(input$categoryInput, levels = levels(df$product_category)),
    discount_percent = as.numeric(input$discountInput),
    shipping_cost    = as.numeric(input$shippingInput),
    rating           = as.numeric(input$ratingInput)
  )
  
  # Predict probability directly in R using tidymodels
  pred <- predict(boost_fit, new_data = new_order, type = "prob")
  prob <- pred$.pred_Yes
  prob_pct <- round(prob * 100, 1)
  
  # Dynamic color threshold matching original UI
  bar_color <- if (prob > 0.5) "#c0392b" else "#2e7d32"
  
  tagList(
    div(style = "font-size: 1.4em; font-weight: bold; margin-bottom: 6px;", 
        paste0(prob_pct, "%")),
    div(style = "width: 100%; max-width: 500px; background-color: #eee; border-radius: 6px; height: 20px;",
        div(style = sprintf("width: %s%%; background-color: %s; height: 100%%; border-radius: 6px; transition: width 0.3s;", 
                            prob_pct, bar_color)))
  )
})
```

## 2.3 Experiment Log

| Scenario | Category | Discount % | Shipping $ | Rating | Predicted P(Return) |
|------------|------------|------------|------------|------------|------------|
| Low risk | Electronics | 5 | 5 | 4.8 | *[fill in]* |
| High discount, low rating | Fashion | 60 | 10 | 1.8 | *[fill in]* |
| High shipping cost | Footwear | 20 | 35 | 3.5 | *[fill in]* |
| Baseline / average order | Accessories | 20 | 8 | 3.5 | *[fill in]* |

**Observations:** *[Does the gap between the low-risk and high-discount/ low-rating scenarios support the RO3 hypothesis that high discount + low rating drives returns?]*

## 2.4 Deployed Model on the Web

Published via shinyapps.io

Published via Shiny 
