---
title: "Reflection on Trees with Group Project Data"
subtitle: "Analytics Objective: Predict whether a purchased item will be returned"
author: "Brian V"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: 1
    toc-location: right-body
    toc-title: "Contents"
    number-sections: true
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    highlight-style: github
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
  freeze: true  # Turn back on after the notebook renders cleanly
---

> This report is my Module 9 reflection on decision trees, applied to my
> Research Objective 3 (RO3) data from our group's Shopify returns-prediction
> project. Instructor prompts are reproduced in block quotes; my responses
> follow directly underneath each one, per the submission instructions.

# Analytics Objective (AO)

> **Prompt:** State your AO (Analytics Objectives) for your project. If your
> AO didn't involve machine learning (supervised or unsupervised learning),
> please revise it so that the objective is appropriately stated for a
> machine learning task. Further state how you will address the objectives.
> Specifically describe: (a) supervised vs. unsupervised learning, (b) the
> variables to be included and their roles (outcome vs. features), and (c)
> the name of the broader category of machine learning methods (regression
> vs. classification).

**AO:** From the perspective of the business, we want to know: *before an
order ships, how likely is this customer's purchase to come back as a
return?* Answering that lets a retailer intervene proactively — flagging
high-risk orders for extra quality checks, adjusting discount policy in
categories that are prone to returns, or setting customer expectations up
front — rather than only reacting to returns after they happen. Framed as an
analytics objective: **predict whether a purchased item will be returned
(`is_returned`) using product category, discount percentage, shipping cost,
and customer rating, in order to identify the strongest drivers of return
risk and estimate their downstream effect on net profit.** This AO already
involved supervised learning in my Module 8 submission, so it did not need
to be revised for this module — I'm simply extending the same objective with
a new class of model (decision trees) instead of stating it from scratch.

- **Learning type:** Supervised learning. `is_returned` is a known, labeled
  outcome recorded for every one of the 60,000 historical orders, so the
  models are trained on examples that already have a "right answer."
- **Variables and roles:**

| Variable | Role | Type | Description |
|:---|:---|:---|:---|
| `is_returned` | Outcome (DV) | Binary factor (No/Yes) | Whether the order was returned |
| `product_category` | Feature (IV) | Categorical | Product line purchased |
| `discount_percent` | Feature (IV) | Numeric | % discount applied to the order |
| `shipping_cost` | Feature (IV) | Numeric | Logistics cost absorbed on the order |
| `rating` | Feature (IV) | Numeric | Post-purchase satisfaction score (1-5) |

- **Broader ML category:** **Classification.** The outcome is categorical
  (returned vs. not returned) rather than a continuous number, so this is a
  binary classification task, not a regression task — the same framing I
  used in Module 8 with logistic regression, LDA, QDA, and KNN.

# Choosing a Decision Tree Method

> **Prompt:** Choose either of the two decision tree methods for your chosen
> machine learning task (i.e., regression vs. classification — help me
> choose): Classification Tree or Regression Tree.

Since `is_returned` is a binary (Yes/No) outcome and not a continuous
number, this AO calls for a **classification tree**, not a regression tree.
Regression trees split the feature space to minimize the *sum of squared
residuals* of a continuous target (e.g., predicting `profit` in dollars);
classification trees instead split to minimize a measure of **node
impurity** (Gini index or entropy) so that each resulting leaf is as
"pure" as possible in terms of the class label. Because I want a predicted
class (or predicted probability of return) rather than a predicted dollar
amount, the classification tree is the correct tool here, and it keeps this
module consistent with the classification framing I already established for
RO3 in Module 8.

::: {.callout-tip title="Why a tree at all, alongside logistic regression?"}
Logistic regression assumes a linear relationship (on the log-odds scale)
between each predictor and the outcome. A classification tree makes no such
assumption — it can automatically capture an **interaction** like "high
discount *and* low rating *and* Fashion category" without me having to
specify that interaction term by hand. That is exactly the kind of pattern
our RO3 hypothesis proposed, so trees are a natural next step.
:::

# Libraries and Data

```{r}
library(tidyverse)   # dplyr, ggplot2, lubridate, readr
library(tidymodels)  # rsample, recipes, parsnip, workflows, yardstick, tune
library(rpart.plot)  # visualizing a single decision tree
library(baguette)    # bag_tree()
library(ranger)      # random forest engine
library(xgboost)     # boosted trees engine
library(doParallel)  # parallel backend to speed up CV

tidymodels_prefer()

# Use multiple cores for cross-validation (leaves 1 core free)
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

df <- read_csv("shopify_sales_dataset_ml_eda.csv")
glimpse(df)
```

# Data Wrangling

```{r}
df <- df %>%
  mutate(
    order_date = ymd(order_date),
    net_margin = (profit / revenue) * 100,
    is_returned = factor(is_returned, levels = c(0, 1), labels = c("No", "Yes")),
    product_category = as.factor(product_category),
    customer_country = as.factor(customer_country)
  )

# Check for missing data
sum(is.na(df))

# Class balance -- returns remain the minority class
df %>%
  count(is_returned) %>%
  mutate(pct = round(n / sum(n) * 100, 1))
```

::: {.callout-note title="Class imbalance, revisited"}
As in Module 8, roughly 15% of orders are returned.[^imbalance] All splits
below are stratified on `is_returned`, and I evaluate models on ROC AUC,
sensitivity, and specificity together — not accuracy alone — since a model
that predicts "No" for every order would still look ~85% accurate while
being operationally useless.
:::

[^imbalance]: A tree-based model can actually handle imbalance a little more
    gracefully than logistic regression in one respect: because splits are
    chosen to reduce impurity locally, a tree can carve out a small, pure
    "high risk" leaf even when the overall return rate is low, whereas a
    single global logistic-regression boundary has to spread its
    sensitivity across the whole feature space.

# Machine Learning Process

## Method Choice: Why a Decision Tree First

I start with a single classification tree (via the `rpart` engine) as the
baseline tree-based model for three reasons: (1) it is the most
**interpretable** non-linear model available — the fitted tree can be drawn
and read like a flowchart of business rules ("if `discount_percent` > 35 and
`rating` < 3, predict Return"); (2) it requires almost **no preprocessing**
— no dummy encoding or normalization is needed, since `rpart` splits factors
and numeric variables directly; and (3) it gives me a natural baseline to
compare against **tree ensembles** (bagging, random forest, boosting), which
should outperform a single tree if the extra variance-reduction or
bias-reduction machinery is actually buying us anything on this dataset.

## Data Splitting

```{r}
set.seed(123)
data_split <- initial_split(df, prop = 0.75, strata = is_returned)
train_data <- training(data_split)
test_data  <- testing(data_split)
```

## Recipe Creation

Tree-based models (`rpart`, `ranger`) split on raw factor levels and raw
numeric values directly, so — unlike the logistic/LDA/QDA/KNN recipe from
Module 8 — this recipe skips `step_dummy()` and `step_normalize()`. The one
exception is **xgboost**, which needs a fully numeric input matrix, so I
build a second recipe with dummy coding for that engine only.

```{r}
tree_recipe <- recipe(is_returned ~ product_category + discount_percent +
                         shipping_cost + rating, data = train_data) %>%
  step_novel(all_nominal_predictors()) %>%
  step_zv(all_predictors())

boost_recipe <- tree_recipe %>%
  step_dummy(all_nominal_predictors())
```

## Model Specification

```{r}
# Tree counts and times reduced significantly below to drastically improve Positron render speeds
tree_spec <- decision_tree(cost_complexity = 0.01, tree_depth = 30, min_n = 20) %>%
  set_engine("rpart") %>%
  set_mode("classification")

bag_spec <- bag_tree() %>%
  set_engine("rpart", times = 5) %>%           # 10 -> 5 for faster rendering
  set_mode("classification")

rf_spec <- rand_forest(trees = 50, mtry = 3, min_n = 10) %>%  # 200 -> 50 trees for faster rendering
  set_engine("ranger", importance = "impurity", num.threads = parallel::detectCores() - 1) %>%
  set_mode("classification")

boost_spec <- boost_tree(trees = 50, tree_depth = 6, learn_rate = 0.05) %>%  # 200 -> 50 trees for faster rendering
  set_engine("xgboost", nthread = parallel::detectCores() - 1) %>%
  set_mode("classification")
```

## Workflows and Fitting

```{r}
tree_wf  <- workflow() %>% add_recipe(tree_recipe)  %>% add_model(tree_spec)
bag_wf   <- workflow() %>% add_recipe(tree_recipe)  %>% add_model(bag_spec)
rf_wf    <- workflow() %>% add_recipe(tree_recipe)  %>% add_model(rf_spec)
boost_wf <- workflow() %>% add_recipe(boost_recipe) %>% add_model(boost_spec)

tree_fit  <- fit(tree_wf,  data = train_data)
bag_fit   <- fit(bag_wf,   data = train_data)
rf_fit    <- fit(rf_wf,    data = train_data)
boost_fit <- fit(boost_wf, data = train_data)
```

### Visualizing the Single Tree {#sec-tree-plot}

```{r}
#| label: fig-single-tree
#| fig-cap: "Fitted classification tree for predicting is_returned"
rpart.plot(extract_fit_engine(tree_fit), roundint = FALSE, type = 3,
           extra = 104, box.palette = "Blues")
```

@fig-single-tree shows the actual decision rules the tree learned. Reading
the top split tells us which single variable most reduces impurity first —
this is usually `discount_percent` or `rating`, consistent with our RO3
hypothesis that heavy discounting combined with low satisfaction is the
main driver of returns.

## Model Evaluation (Test Set)

```{r}
model_fits <- list(
  "Single Tree"   = tree_fit,
  "Bagged Trees"  = bag_fit,
  "Random Forest" = rf_fit,
  "Boosted Trees" = boost_fit
)

test_preds <- map(model_fits, augment, new_data = test_data)

test_metrics <- map_dfr(test_preds, \(preds) {
  bind_rows(
    roc_auc(preds, truth = is_returned, .pred_Yes, event_level = "second"),
    accuracy(preds, truth = is_returned, estimate = .pred_class),
    sens(preds, truth = is_returned, estimate = .pred_class, event_level = "second"),
    spec(preds, truth = is_returned, estimate = .pred_class, event_level = "second")
  )
}, .id = "model")

test_metrics %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)
```

::: {.panel-tabset}
## Single Tree
```{r}
conf_mat(test_preds[["Single Tree"]], truth = is_returned, estimate = .pred_class)
```

## Bagged Trees
```{r}
conf_mat(test_preds[["Bagged Trees"]], truth = is_returned, estimate = .pred_class)
```

## Random Forest
```{r}
conf_mat(test_preds[["Random Forest"]], truth = is_returned, estimate = .pred_class)
```

## Boosted Trees
```{r}
conf_mat(test_preds[["Boosted Trees"]], truth = is_returned, estimate = .pred_class)
```
:::

**Interpretation:** *[Fill in with your actual rendered numbers.]* In
general, expect the single tree to have the lowest AUC of the four — it is
prone to high variance and tends to overfit its training splits. Bagging
should reduce that variance by averaging many bootstrapped trees; the random
forest typically improves on plain bagging further by decorrelating trees
through random feature subsetting at each split; and boosting often edges
out the random forest on AUC because it builds trees sequentially to correct
prior errors, at the cost of being the slowest to fit and the most sensitive
to overfitting if `trees` or `learn_rate` are not tuned carefully.

## Alternative Methods Tried

The four models above — a single classification tree, bagged trees, a
random forest, and boosted trees — *are* the alternative methods used to try
to improve on the baseline single tree, per the assignment prompt. Bagging
and random forests both attack the single tree's **high variance** problem
by averaging over many trees fit to resampled data (random forests add
random feature subsetting to further decorrelate those trees), while
boosting instead attacks **bias**, building shallow trees sequentially so
each new tree focuses on the previous ensemble's mistakes.

## Cross-Validation Comparison

```{r}
set.seed(123)
folds <- vfold_cv(train_data, v = 3, strata = is_returned)  # 5 -> 3 folds for much faster rendering

wf_list <- list(
  "Single Tree"   = tree_wf,
  "Bagged Trees"  = bag_wf,
  "Random Forest" = rf_wf,
  "Boosted Trees" = boost_wf
)

cv_metrics <- metric_set(roc_auc, accuracy, sens, spec)
cv_control <- control_resamples(allow_par = TRUE)  # let doParallel spread the fits

cv_results <- map_dfr(wf_list, \(wf) {
  fit_resamples(wf, resamples = folds, metrics = cv_metrics, control = cv_control) %>%
    collect_metrics()
}, .id = "model")

cv_results %>%
  select(model, .metric, mean, std_err) %>%
  pivot_wider(names_from = .metric, values_from = c(mean, std_err))
```

**Interpretation:** *[Fill in once rendered.]* Compare the `mean` AUC
column across models and check whether the `std_err` bands overlap. If the
random forest and boosted trees are within roughly one standard error of
each other, that's evidence the two ensembles are performing comparably and
the simpler-to-tune one (usually the random forest) is the more practical
choice for deployment. If the single tree's CV AUC is noticeably below the
ensembles, that confirms the single tree was overfitting to idiosyncrasies
of the training split rather than learning generalizable rules — the
expected result, since averaging/boosting exist precisely to fix that
weakness.

## Feature Importance

```{r}
#| label: fig-rf-importance
#| fig-cap: "Random forest feature importance (impurity-based)"

# Extracting feature importance using base ggplot2 instead of the vip package
rf_obj <- extract_fit_engine(rf_fit)

importance_df <- tibble(
  feature = names(rf_obj$variable.importance),
  importance = rf_obj$variable.importance
) %>%
  arrange(desc(importance)) %>%
  slice_head(n = 10)

ggplot(importance_df, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Feature Importance for Predicting is_returned",
       x = "Feature", y = "Importance (Impurity)") +
  theme_minimal()
```

**Interpretation:** *[Fill in once rendered.]* Compare this ranking to the
Module 8 logistic-regression coefficients and to our original RO3
hypothesis (heavy discounting + low rating, concentrated in Fashion and
Footwear, drives returns). If `discount_percent` and `rating` again come out
on top, that's convergent evidence across two very different model families
— logistic regression's linear odds ratios and the random forest's
impurity-based importance — which strengthens the case that these two
variables are genuine drivers of return risk rather than an artifact of one
particular model.

# Decision Tree Pruning

> **Prompt:** Using the `fit_tree_cp()` function from Section 4.5 (or write
> your own), fit decision trees across `cost_complexity` values of
> `c(0.0001, 0.001, 0.005, 0.01, 0.05, 0.10)`. Plot train vs. test AUC across
> all `cp` values. At what `cp` value does test AUC peak? What happens to
> train AUC at that same value? At the optimal `cp`, how many leaf nodes
> does the tree have? In plain English, explain to a marketing manager why a
> tree with 200 leaf nodes might perform worse on new customers than a tree
> with 15 leaf nodes, even though it was more accurate on the training data.

## Fitting Trees Across `cost_complexity` Values

```{r}
fit_tree_cp <- function(cp, train, test, recipe) {
  spec <- decision_tree(cost_complexity = cp) %>%
    set_engine("rpart") %>%
    set_mode("classification")

  wf <- workflow() %>% add_recipe(recipe) %>% add_model(spec)
  fitted <- fit(wf, data = train)

  train_auc <- augment(fitted, new_data = train) %>%
    roc_auc(truth = is_returned, .pred_Yes, event_level = "second") %>%
    pull(.estimate)

  test_auc <- augment(fitted, new_data = test) %>%
    roc_auc(truth = is_returned, .pred_Yes, event_level = "second") %>%
    pull(.estimate)

  n_leaves <- fitted %>%
    extract_fit_engine() %>%
    pluck("frame") %>%
    filter(var == "<leaf>") %>%
    nrow()

  tibble(cost_complexity = cp, train_auc = train_auc,
         test_auc = test_auc, n_leaves = n_leaves)
}

cp_values <- c(0.0001, 0.001, 0.005, 0.01, 0.05, 0.10)

cp_results <- map_dfr(cp_values, fit_tree_cp,
                       train = train_data, test = test_data,
                       recipe = tree_recipe)

cp_results
```

## Train vs. Test AUC Across `cp` Values

```{r}
#| label: fig-cp-auc
#| fig-cap: "Train vs. test AUC across cost_complexity (cp) values"
cp_results %>%
  pivot_longer(cols = c(train_auc, test_auc),
               names_to = "set", values_to = "auc") %>%
  mutate(set = recode(set, train_auc = "Train", test_auc = "Test")) %>%
  ggplot(aes(x = cost_complexity, y = auc, color = set)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(x = "cost_complexity (log scale)", y = "ROC AUC", color = NULL,
       title = "Train vs. Test AUC Across Pruning Levels") +
  theme_minimal()
```

@fig-cp-auc typically shows the familiar pruning pattern for a single tree:
**train AUC decreases monotonically as `cp` grows** (a larger `cp` forces a
smaller, more heavily pruned tree, which necessarily fits the training data
less closely), while **test AUC rises to a peak at some intermediate `cp`
and then falls off** once the tree becomes too shallow to capture real
signal. *[Report your actual peak `cp` value and both AUCs here once
rendered — e.g., "test AUC peaks at cp = ___ (test AUC = ___), while train
AUC at that same cp is ___, only slightly higher than the peak test AUC."]*

## Leaf Nodes at the Optimal `cp`

```{r}
best_cp <- cp_results %>% slice_max(test_auc, n = 1) %>% pull(cost_complexity)
best_cp

cp_results %>% filter(cost_complexity == best_cp) %>% pull(n_leaves)
```

*[State the resulting leaf count here, e.g., "At the optimal cp of ___, the
pruned tree has ___ leaf nodes — far fewer than the ~200+ leaves an
essentially unpruned tree (cp = 0.0001) would produce."]*

## Explaining Overfitting to a Marketing Manager

> A tree with 200 leaf nodes has essentially memorized the training
> customers rather than learning general rules about *why* people return
> items. Each of those 200 tiny leaves might represent a rule as specific as
> "customers in Germany who bought Footwear with exactly a 27% discount and
> a 3.4 rating on a Tuesday" — a pattern that fit our historical data almost
> perfectly by chance, but that has no real predictive power for a *new*
> customer who doesn't match that razor-thin combination. A tree with 15
> leaf nodes, by contrast, is forced to group customers using broader,
> more stable patterns — "high discount and low rating," for instance —
> that are far more likely to hold up for future orders. So while the
> 200-leaf tree looks more accurate when we check it against the same data
> it was trained on, the 15-leaf tree usually wins on genuinely new orders,
> because it learned the underlying *signal* in customer behavior instead of
> memorizing *noise* specific to our training sample. This is exactly why
> we don't just pick the tree with the highest training accuracy — we pick
> the one with the best **test** performance.

::: {.callout-important title="Business takeaway"}
Overfitting isn't just a modeling technicality — an over-fit return-risk
model would flag the wrong customers, wasting review-team effort on
one-off historical quirks instead of catching the genuinely risky
discount/rating combinations that recur order after order.
:::

```{r}
# Shut down the parallel cluster now that all model fitting is done
stopCluster(cl)
registerDoSEQ()
```



## Appendix

Github Repo: https://github.com/brianvuong1/r-project-git

