---
title: "M05: Reflection on Ch09 (Evaluation)"
author: "Brian V"
date: today
format:

  html:
    toc: true
    toc-depth: 4
    toc-expand: 1
    toc-location: right-body
    toc-title: "Contents"
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    highlight-style: github
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
  freeze: true
---



# Setup

```{r setup}
library(tidymodels) # umbrella: rsample, recipes, parsnip, workflows, yardstick
library(tidyverse)
library(gt) # pretty tables
library(patchwork) # combine plots

tidymodels_prefer() # resolve function-name conflicts
set.seed(2026)
```

------------------------------------------------------------------------

# The Marketing Dataset

We return to our synthetic **direct-mail** dataset from Week 2. Each row is one prospect; the outcome is whether they **responded** (bought).

```{r load-data}
# ── Synthetic direct-mail dataset (same generator as Week 2 / Week 7) ─────────
set.seed(123)
n <- 3000

mail_data <- tibble(
  customer_id = paste0("C", str_pad(1:n, 5, pad = "0")),
  age = round(rnorm(n, mean = 45, sd = 12)),
  income = round(rlnorm(n, meanlog = 10.8, sdlog = 0.6)), # right-skewed
  recency_days = round(rexp(n, rate = 1 / 60)), # days since last purchase
  freq_12mo = rpois(n, lambda = 3), # purchases in 12 months
  avg_order_amt = round(rlnorm(n, meanlog = 4.2, sdlog = 0.5), 2),
  channel = sample(
    c("email", "direct_mail", "digital"),
    n,
    replace = TRUE,
    prob = c(0.5, 0.3, 0.2)
  ),
  region = sample(
    c("West", "South", "Midwest", "Northeast"),
    n,
    replace = TRUE
  ),
  loyalty_tier = factor(
    sample(
      c("Bronze", "Silver", "Gold", "Platinum"),
      n,
      replace = TRUE,
      prob = c(0.4, 0.3, 0.2, 0.1)
    ),
    levels = c("Bronze", "Silver", "Gold", "Platinum")
  ),
  responded = rbinom(
    n,
    1,
    prob = plogis(
      -3 +
        0.02 * (age - 45) +
        0.3 * log(income / 50000) +
        0.1 * freq_12mo -
        0.005 * recency_days
    )
  )
) |>
  mutate(
    income = if_else(runif(n) < 0.08, NA_real_, income),
    avg_order_amt = if_else(runif(n) < 0.05, NA_real_, avg_order_amt),
    responded = factor(responded, levels = c(1, 0), labels = c("yes", "no"))
  )
# introduce ~8 % missingness in income and avg_order_amt

glimpse(mail_data)
```

We also derive a **continuous** target — `order_value_next` — so we have a realistic stand-in for regression metrics in Section 1. Think of this as "expected next-order revenue," something a marketer might forecast for budget planning.

```{r derive-regression-target}
mail_data <- mail_data |>
  mutate(
    order_value_next = round(
      pmax(
        0,
        15 +
          0.0006 * coalesce(income, median(income, na.rm = TRUE)) +
          4 * freq_12mo -
          0.05 * recency_days +
          rnorm(n, 0, 20)
      ),
      2
    )
  )

mail_data |> select(customer_id, order_value_next) |> head()
```

::: callout-note
**Marketing context.** `responded` (yes/no) is our **classification** target — will this prospect respond to the next campaign? `order_value_next` is our **regression** target — how much will they spend on their next order? Both are common marketing prediction tasks, and TMWR Chapter 9 covers metrics for each.
:::

------------------------------------------------------------------------

# Train/Test Split and a Reference Model

Before we can evaluate anything, we need a fitted model. We reuse the same split-then-recipe workflow from Week 2.

```{r split}
set.seed(617)
mail_split <- initial_split(mail_data, prop = 0.80, strata = responded)
mail_train <- training(mail_split)
mail_test <- testing(mail_split)

cat("Training rows:", nrow(mail_train), "\n")
cat("Test rows    :", nrow(mail_test), "\n")
```

```{r recipe-model}
mail_rec <- recipe(
  responded ~ age +
    income +
    recency_days +
    freq_12mo +
    avg_order_amt +
    channel +
    region +
    loyalty_tier,
  data = mail_train
) |>
  step_impute_median(all_numeric_predictors()) |>
  step_log(income, avg_order_amt, recency_days, base = 10, offset = 1) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

lr_spec <- logistic_reg(penalty = 0.01, mixture = 1) |> # Lasso
  set_engine("glmnet") |>
  set_mode("classification")

mail_wf <- workflow() |>
  add_recipe(mail_rec) |>
  add_model(lr_spec)

mail_fit <- fit(mail_wf, data = mail_train)
```

```{r predict-test}
test_preds <- augment(mail_fit, new_data = mail_test)
test_preds |>
  select(customer_id, responded, .pred_class, .pred_yes, .pred_no) |>
  head()
```

------------------------------------------------------------------------

# **Part 1** — Judging Model Effectiveness (TMWR Ch. 9)

-   A fitted model is only useful if we can say, in concrete numeric terms, *how good* its predictions are. 

-   `yardstick` is the tidymodels package for this — every metric function takes `truth` and an `estimate`/probability column and returns a tidy one-row tibble.

## Why Metrics, Not Just "It Fit"

::: callout-important
**There is no single "best" metric.** The right metric depends on the business question. For a churn model, missing a true churner (low **sensitivity**) might be far more costly than flagging a loyal customer by mistake (low **specificity**). Always pick metrics that match the marketing decision the model supports.
:::

## Regression Metrics

We don't have a regression *model* yet in this notebook, so we'll fit one quickly — predicting `order_value_next` from a few predictors — purely to demonstrate the metrics.

```{r reg-model}
reg_spec <- linear_reg() |>
  set_engine("lm")

reg_rec <- recipe(
  order_value_next ~ income + freq_12mo + recency_days,
  data = mail_train
) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors())

reg_wf <- workflow() |> add_recipe(reg_rec) |> add_model(reg_spec)
reg_fit <- fit(reg_wf, data = mail_train)

reg_preds <- augment(reg_fit, new_data = mail_test)
reg_preds |> select(customer_id, order_value_next, .pred, .resid) |> head()
```

```{r reg-metrics}

# remove the conflicting object, recreate a regression metric-set, then call it
rm(metric_set)
reg_metrics <- yardstick::metric_set(rmse, rsq, mae)
reg_preds |> reg_metrics(truth = order_value_next, estimate = .pred)


reg_metrics <- metric_set(rmse, rsq, mae)

reg_preds |>
  reg_metrics(truth = order_value_next, estimate = .pred)
```

### **RMSE** (root mean squared error): 

Average prediction error, in the *original units* (dollars here). Penalizes large errors more heavily because errors are squared before averaging.

::: {.callout-note title="RMSE Formula"}
$$
RMSE = \sqrt{\frac{1}{n}\sum_{i=1}^{n}(y_i - \hat{y}_i)^2}
$$

- Where 

  - $n$ is the number of observations, 
  - $y_i$ is the actual value, and 
  - $\hat{y}_i$ is the predicted value. 
  - $(y_i - \hat{y}_i)$ = the residual (prediction error) for observation $i$

- RMSE is in the **same units as the outcome variable** — dollars, in our case — making it directly interpretable.
:::

### **R² (rsq)**:

- Proportion of variance in the outcome explained by the model. 

- Ranges roughly 0–1; higher is better. 

- Useful for communicating "how much of next-order value our model explains" to stakeholders.

::: {.callout-note title="R-Squared formula"}
$$
R^2 = 1 - \frac{\sum_{i=1}^{n}(y_i - \hat{y}_i)^2}{\sum_{i=1}^{n}(y_i - \bar{y})^2}
$$

Where:

- $y_i$ = actual value for observation $i$
- $\hat{y}_i$ = predicted value for observation $i$
- $\bar{y}$ = mean of all actual values
- $\sum_{i=1}^{n}(y_i - \hat{y}_i)^2$ = total unexplained variation (residual sum of squares)
- $\sum_{i=1}^{n}(y_i - \bar{y})^2$ = total variation in the outcome (total sum of squares)
:::

### **MAE** (mean absolute error): 

- Average absolute error, less sensitive to outliers than RMSE. 

- If a few extreme spenders are skewing RMSE, compare it against MAE.

::: {.callout-note title="MAE Formula"}
$$
MAE = \frac{1}{n}\sum_{i=1}^{n}|y_i - \hat{y}_i|
$$

Where:

- $n$ = number of observations
- $y_i$ = actual value for observation $i$
- $\hat{y}_i$ = predicted value for observation $i$
- $|y_i - \hat{y}_i|$ = absolute prediction error for observation $i$
:::

```{r reg-plot}
#| fig-height: 4.5
#| fig-width: 6
ggplot(reg_preds, aes(x = order_value_next, y = .pred)) +
  geom_point(alpha = 0.3, color = "#2E5266") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "tomato") +
  labs(
    title = "Predicted vs. Actual Next-Order Value",
    subtitle = "Dashed line = perfect prediction",
    x = "Actual order value ($)",
    y = "Predicted order value ($)"
  ) +
  theme_minimal(base_size = 12)
```

## Binary Classification Metrics

Back to our `responded` (yes/no) model. The starting point for classification metrics is the **confusion matrix**: a 2×2 table of predicted vs. actual class.

```{r conf-mat}
conf_mat(test_preds, truth = responded, estimate = .pred_class)
```

```{r conf-mat-plot}
#| fig-height: 4
#| fig-width: 4.5
conf_mat(test_preds, truth = responded, estimate = .pred_class) |>
  autoplot(type = "heatmap")
```

::: {.callout-note title="From Confusion Matrix to Metrics"}
| Metric | Formula (informal) | Marketing question it answers |
|----|----|----|
| **Sensitivity** (recall) | TP / (TP + FN) | Of customers who *would* respond, how many did we catch? |
| **Specificity** | TN / (TN + FP) | Of customers who *wouldn't* respond, how many did we correctly skip? |
| **Precision (PPV)** | TP / (TP + FP) | Of customers we targeted, how many actually responded? |
| **Accuracy** | (TP + TN) / Total | Overall, how often is the model right? *(misleading with rare classes — see warning below)* |
:::

### Sensitivity (Recall)

::: {.callout-note title="Details"}
$$
\text{Sensitivity} = \frac{TP}{TP + FN}
$$

Where:

- $TP$ = true positives — customers who responded and were correctly predicted to respond
- $FN$ = false negatives — customers who responded but were incorrectly predicted not to respond
- $TP + FN$ = total number of actual responders
:::

### Specificity

::: {.callout-note title="Details"}
$$
\text{Specificity} = \frac{TN}{TN + FP}
$$

Where:

- $TN$ = true negatives — customers who did not respond and were correctly predicted not to respond
- $FP$ = false positives — customers who did not respond but were incorrectly predicted to respond
- $TN + FP$ = total number of actual non-responders
:::

### Precision (PPV)

::: {.callout-note title="Details"}
$$
\text{Precision} = \frac{TP}{TP + FP}
$$

Where:

- $TP$ = true positives — correctly predicted responders
- $FP$ = false positives — non-responders incorrectly predicted as responders
- $TP + FP$ = total number of customers predicted to respond
:::

### Accuracy

::: {.callout-note title="Details"}
$$
\text{Accuracy} = \frac{TP + TN}{TP + TN + FP + FN}
$$

Where:

- $TP$ = true positives
- $TN$ = true negatives
- $FP$ = false positives
- $FN$ = false negatives
- $TP + TN + FP + FN$ = total number of observations
:::

### F1 Score

::: {.callout-note title="Details"}
$$
F_1 = 2 \times \frac{\text{Precision} \times \text{Sensitivity}}{\text{Precision} + \text{Sensitivity}}
$$

Where:

- $\text{Precision} = \frac{TP}{TP + FP}$
- $\text{Sensitivity} = \frac{TP}{TP + FN}$
- $F_1$ is the harmonic mean of precision and sensitivity — useful when both matter equally and the classes are imbalanced
:::

```{r class-metrics}
class_metrics <- metric_set(accuracy, sens, spec, precision, f_meas)

test_preds |>
  class_metrics(truth = responded, estimate = .pred_class)
```

::: callout-warning
**Accuracy is dangerous with imbalanced classes.** Direct-mail response rates are often only 3–8%. A model that predicts "no" for *everyone* could still score 95%+ accuracy while being completely useless for targeting. Always check sensitivity/specificity alongside accuracy — this is exactly the kind of trap a stakeholder unfamiliar with the data could fall into.
:::

### ROC Curve and AUC

Sensitivity and specificity above were computed at a single classification threshold (0.5 by default). The **ROC curve** shows the sensitivity/specificity trade-off across *every possible* threshold, and **AUC** summarizes the whole curve in one number.

```{r roc-curve}
#| fig-height: 4.5
#| fig-width: 5
roc_data <- test_preds |>
  roc_curve(truth = responded, .pred_yes)

auc_val <- test_preds |>
  roc_auc(truth = responded, .pred_yes) |>
  pull(.estimate)
```

```{r roc-curve-plot}
autoplot(roc_data) +
  annotate(
    "text",
    x = 0.6,
    y = 0.2,
    label = paste0("AUC = ", round(auc_val, 3)),
    size = 5,
    color = "#2E5266",
    fontface = "bold"
  ) +
  labs(
    title = "ROC Curve — Direct-Mail Response Model",
    subtitle = "Closer to the top-left corner = better discrimination"
  ) +
  theme_minimal(base_size = 12)
```

::: callout-tip
**Rule of thumb.** AUC = 0.5 means the model is no better than a coin flip. AUC = 1.0 is a perfect classifier (rare in practice). For direct-mail targeting, AUC \> 0.65 is often usable; \> 0.75 is quite good.
:::

## Multiclass Classification Metrics

`loyalty_tier` (Bronze/Silver/Gold/Platinum) is a 4-level outcome. While we won't build a full multiclass model here, it's worth knowing how the binary metrics above generalize.

```{r multiclass-demo}
# Demonstration only: predict loyalty_tier from age + income as a toy multiclass model
multi_rec <- recipe(loyalty_tier ~ age + income, data = mail_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors())

multi_spec <- multinom_reg() |> set_engine("nnet")

multi_wf <- workflow() |> add_recipe(multi_rec) |> add_model(multi_spec)
multi_fit <- fit(multi_wf, data = mail_train)

multi_preds <- augment(multi_fit, new_data = mail_test)

multi_preds |>
  metric_set(accuracy, sens)(truth = loyalty_tier, estimate = .pred_class)

multi_preds |>
  roc_auc(
    truth = loyalty_tier,
    .pred_Bronze,
    .pred_Silver,
    .pred_Gold,
    .pred_Platinum
  ) |>
  pull(.estimate)
```

::: {.callout-note title="How Binary Metrics Generalize to Multiclass"}
With more than two classes, `yardstick` computes a metric *per class* (treating each as "this class" vs. "all others"), then **averages** across classes. The default is **macro-averaging** — each class counts equally, regardless of how many observations it has. This matters for `loyalty_tier`: Platinum customers are rare, but macro-averaging keeps them from being drowned out by the much larger Bronze group.
:::

### Estimator Argument (Optional)

> Refer to Chapter 10
```{r multiclass-estimaor}
multi_preds |>
  metric_set(accuracy, sens)(
    truth = loyalty_tier,
    estimate = .pred_class,
    estimator = "macro"
  )

multi_preds |>
  metric_set(accuracy, sens)(
    truth = loyalty_tier,
    estimate = .pred_class,
    estimator = "macro_weighted"
  )

multi_preds |>
  metric_set(accuracy, sens)(
    truth = loyalty_tier,
    estimate = .pred_class,
    estimator = "micro"
  )

multi_preds |>
  roc_auc(
    truth = loyalty_tier,
    .pred_Bronze,
    .pred_Silver,
    .pred_Gold,
    .pred_Platinum,
    estimator = "macro_weighted"
  )
```

------------------------------------------------------------------------

# **Part 2** — Resampling for Evaluating Performance (TMWR Ch. 10)

The metrics above were all computed on a **single** test set. That gives us *a* number — but how much would that number change if we'd happened to split the data differently? Resampling answers this question.

## The Resubstitution Problem

What happens if we evaluate the model on the **same data it was trained on**?

```{r resubstitution}
train_preds <- augment(mail_fit, new_data = mail_train)

bind_rows(
  train_preds |>
    class_metrics(truth = responded, estimate = .pred_class) |>
    mutate(data = "training (resubstitution)"),
  test_preds |>
    class_metrics(truth = responded, estimate = .pred_class) |>
    mutate(data = "test (held out)")
) |>
  select(data, .metric, .estimate) |>
  pivot_wider(names_from = data, values_from = .estimate)
```

::: callout-important
**Resubstitution is optimistic.** The model has already "seen" the training data and partially memorized its noise, so performance measured there is inflated relative to how the model will do on genuinely new prospects. This gap is usually small for simple models like **regularized logistic regression**, but it can be large for flexible models (deep trees, boosted ensembles) — which is exactly why we need a principled way to estimate out-of-sample performance *without* permanently sacrificing data to a single test set.
:::

## Why Not Just Use the Test Set Repeatedly?

The test set should be touched **once**, at the very end, to report final performance. If we use it over and over while comparing models or tuning hyperparameters, we implicitly start fitting to it — defeating its purpose. Resampling lets us get reliable performance estimates **using only the training data**, keeping the test set untouched until final evaluation.

## Cross-Validation

The most common resampling scheme is **v-fold (k-fold) cross-validation**: split the training data into *v* roughly equal folds, train on *v − 1* of them, evaluate on the held-out fold, and rotate until every fold has served as the validation set exactly once.

```{r vfold}
set.seed(2024)
mail_folds <- vfold_cv(mail_train, v = 10, strata = responded)
mail_folds
```

```{r vfold-inspect}
# Look at one fold up close
mail_folds$splits[[1]]
```

::: {.callout-note title="What `vfold_cv()` Just Built"}
`mail_folds` is a tibble of 10 `rsplit` objects. Each one knows which rows of `mail_train` are the analysis set (≈ 90%, used for fitting) and which are the assessment set (≈ 10%, used for scoring) for that fold. `strata = responded` keeps the response rate roughly constant across folds — same logic as stratifying the original train/test split in Week 2.
:::

### Visualizing the Fold Structure

```{r fold-viz}
#| fig-height: 4
#| fig-width: 7
fold_viz <- map_dfr(seq_along(mail_folds$splits), function(i) {
  split <- mail_folds$splits[[i]]
  tibble(
    fold = i,
    row = 1:nrow(mail_train),
    role = if_else(
      row %in% split$in_id,
      "Analysis (train)",
      "Assessment (validate)"
    )
  )
})

ggplot(fold_viz, aes(x = row, y = factor(fold), fill = role)) +
  geom_tile() +
  scale_fill_manual(
    values = c(
      "Analysis (train)" = "#A9C0CC",
      "Assessment (validate)" = "#C0504D"
    )
  ) +
  labs(
    title = "10-Fold Cross-Validation Structure",
    subtitle = "Each row of the training data is held out exactly once",
    x = "Training set row index",
    y = "Fold",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Other Resampling Methods (Brief Tour)

```{r other-resampling}
# Repeated CV: run 10-fold CV five times with different random splits
repeated_folds <- vfold_cv(mail_train, v = 10, repeats = 5, strata = responded)
cat("Repeated CV total resamples:", nrow(repeated_folds), "\n")

# Bootstrap: sample WITH replacement, same size as training data
set.seed(2024)
boot_resamples <- bootstraps(mail_train, times = 25, strata = responded)
cat("Bootstrap resamples:", nrow(boot_resamples), "\n")

# Validation set: a single held-out chunk carved from the training data
val_set <- validation_split(mail_train, prop = 0.80, strata = responded)
val_set
```

::: {.callout-note title="Which Resampling Method When?"}
| Method | When to use it | Tradeoff |
|----|----|----|
| **V-fold CV** | Default choice for most marketing datasets | Good balance of bias and computation |
| **Repeated CV** | Smaller datasets where fold-to-fold variability is high | More stable estimates, *v* × *repeats* the compute |
| **Bootstrap** | Want a measure of estimate *uncertainty*, or very small data | Some observations never appear in assessment sets |
| **Validation set** | Very large data where CV is too slow | Only one estimate — less stable than CV |
| **Rolling origin** | Time-ordered data (e.g., monthly campaign cohorts) | Respects temporal order; no shuffling allowed |
:::

## Estimating Performance with `fit_resamples()`

This is where it all comes together: fit the workflow on each fold's analysis set, score it on the assessment set, and average the results.

```{r fit-resamples}
set.seed(2024)
cv_results <- fit_resamples(
  mail_wf,
  resamples = mail_folds,
  metrics = metric_set(accuracy, roc_auc, sens, spec),
  control = control_resamples(save_pred = TRUE)
)

cv_results
```

```{r collect-metrics}
collect_metrics(cv_results)
```

::: {.callout-important title="Comparing the Single Test Set vs. Cross-Validated Estimate"}
```{r compare-estimates}
bind_rows(
  test_preds |>
    metric_set(accuracy, roc_auc)(
      truth = responded,
      estimate = .pred_class,
      .pred_yes
    ) |>
    mutate(source = "Single test set"),
  collect_metrics(cv_results) |>
    filter(.metric %in% c("accuracy", "roc_auc")) |>
    select(.metric, .estimate = mean) |>
    mutate(source = "10-fold CV mean")
) |>
  select(source, .metric, .estimate) |>
  pivot_wider(names_from = source, values_from = .estimate)
```

The single test-set number is **one draw** from a distribution of possible performance estimates. The cross-validated mean averages over 10 different draws, giving a more stable — and more honest — picture of how the model is likely to perform on new prospects. The standard error reported by `collect_metrics()` tells us how much that estimate might wobble.
:::

```{r per-fold-metrics}
collect_metrics(cv_results, summarize = FALSE) |>
  filter(.metric == "roc_auc") |>
  ggplot(aes(x = id, y = .estimate)) +
  geom_col(fill = "#2E5266", alpha = 0.85) +
  geom_hline(
    aes(yintercept = mean(.estimate)),
    linetype = "dashed",
    color = "tomato"
  ) +
  labs(
    title = "ROC AUC by Fold",
    subtitle = "Dashed line = mean across all 10 folds",
    x = "Fold",
    y = "ROC AUC"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Parallel Processing (Conceptual Note)

```{r parallel-note, eval=FALSE}
# Resampling fits the same workflow many times — an embarrassingly parallel task.
# On a multi-core machine, register a parallel backend before fit_resamples():

# install.packages("doParallel")
library(doParallel)
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

cv_results <- fit_resamples(mail_wf, resamples = mail_folds)

stopCluster(cl)
```

::: callout-tip
We don't run this chunk live in class (it depends on your machine's core count), but know that `fit_resamples()`, `tune_grid()`, and friends all parallelize automatically once a backend is registered — useful once your resampling scheme + grid search gets large later in the course.
:::

## Saving Resampled Objects

```{r save-predictions}
# Because we set save_pred = TRUE above, we can pull out every held-out prediction
cv_preds <- collect_predictions(cv_results)
cv_preds |> select(id, .row, responded, .pred_class, .pred_yes) |> head()
```

```{r cv-roc}
#| fig-height: 4.5
#| fig-width: 5
cv_preds |>
  group_by(id) |>
  roc_curve(truth = responded, .pred_yes) |>
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = id)) +
  geom_path(alpha = 0.6, linewidth = 0.8) +
  geom_abline(linetype = "dashed", color = "gray60") +
  labs(
    title = "ROC Curves Across All 10 Folds",
    subtitle = "Spread between lines = fold-to-fold variability",
    color = "Fold"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

::: callout-note
**Why save predictions?** With `save_pred = TRUE`, we get every out-of-fold prediction back, which lets us build diagnostic plots like the multi-fold ROC curve above, inspect specific misclassified prospects, or build custom metrics — all without ever touching the test set.
:::

------------------------------------------------------------------------

# Exercises

## ✏️ Exercise 9.1 — Confusion Matrix and Threshold Tradeoff

**Part A — Reading the Confusion Matrix**

Run `conf_mat(test_preds, truth = responded, estimate = .pred_class)` and
use the output to answer:

1. How many customers actually responded ("yes") and were correctly
   predicted to respond? What is this cell called?
2. How many customers actually responded but were predicted "no"?
   What is this cell called, and what does it cost a marketer?
3. How many customers did not respond but were predicted "yes"?
   What is this cell called, and what does it cost a marketer?
4. In plain English, which type of error is more costly for a
   direct-mail campaign with a $5 mailing cost and $50 average order
   value — a false positive or a false negative? Explain your reasoning.


True Positives: 0 customers. This cell is called True Positives (TP).
False Negatives: 31 customers. This cell is called False Negatives (FN). The cost to a marketer for a false negative is the potential revenue lost from these customers who actually wanted to respond but were not targeted. If the average order value is $50, the cost would be (31 \times 50 = 1550).
False Positives: 0 customers. This cell is called False Positives (FP). The cost to a marketer for a false positive is the cost of mailing to customers who did not respond. If the mailing cost is $5, the cost would be (0 \times 5 = 0).
Costly Error: A false negative is more costly for a direct-mail campaign in this scenario. This is because the marketer misses out on potential sales from customers who would have responded positively, leading to a loss of revenue. In contrast, a false positive incurs only the cost of mailing without generating any revenue, which is less impactful compared to the lost sales from false negatives.

```{r ex9-1a, eval=FALSE}
# Your code here
conf_mat(test_preds, truth = responded, estimate = .pred_class)

```

**Part B — Threshold Tradeoff**

Using `test_preds`, compute `precision` and `sens` at the default 0.5
threshold. Now manually create a new `.pred_class` column using a threshold
of 0.3 instead (classify as "yes" if `.pred_yes > 0.3`). Recompute both
metrics. Which one increased, and why does lowering the threshold have
that effect?

Observations:
Both precision and sensitivity remain at 0 after lowering the threshold to 0.3. This indicates that even with the new threshold, there are still no true positives predicted.

Explanation:
Lowering the threshold generally increases sensitivity because it allows more instances to be classified as positive (in this case, "yes"). However, in this specific scenario, since there were no true positives to begin with, both metrics did not change.
In a typical case, lowering the threshold would lead to an increase in sensitivity (more true positives) but could decrease precision (more false positives). However, since we have no true positives in either case, both metrics remain at zero.

```{r ex9-1b, eval=FALSE}
# Calculate precision and sensitivity at the default threshold of 0.5
TP <- confusion_matrix$True_Positive
FN <- confusion_matrix$False_Negative
FP <- confusion_matrix$False_Positive

# Precision and Sensitivity calculations
precision_default <- ifelse((TP + FP) > 0, TP / (TP + FP), 0)
sensitivity_default <- ifelse((TP + FN) > 0, TP / (TP + FN), 0)

# Create a new .pred_class column using a threshold of 0.3
# Classify as 'yes' if .pred_yes > 0.3
test_preds <- test_preds %>% 
  mutate(.pred_class = ifelse(.pred_yes > 0.3, 'yes', 'no'))

# Recalculate confusion matrix components with the new threshold
new_confusion_matrix <- test_preds %>% 
  summarize(
    True_Positive = sum(responded == 'yes' & .pred_class == 'yes'),
    False_Negative = sum(responded == 'yes' & .pred_class == 'no'),
    False_Positive = sum(responded == 'no' & .pred_class == 'yes')
  )

new_confusion_matrix

precision_default
sensitivity_default

# Recalculate precision and sensitivity with the new threshold
TP_new <- new_confusion_matrix$True_Positive
FN_new <- new_confusion_matrix$False_Negative
FP_new <- new_confusion_matrix$False_Positive

# Precision and Sensitivity calculations for new threshold
precision_new <- ifelse((TP_new + FP_new) > 0, TP_new / (TP_new + FP_new), 0)
sensitivity_new <- ifelse((TP_new + FN_new) > 0, TP_new / (TP_new + FN_new), 0)

precision_new


```

## ✏️ Exercise 9.2 — ROC/AUC Interpretation and Custom `metric_set()`

**Part A — Interpreting AUC**

The notebook computed `roc_auc` for the direct-mail response model.

1. Look up the AUC value from the `roc-curve` chunk. Is it above or
   below 0.65 (the "usable for targeting" threshold from the callout)?
2. In plain English, what does an AUC of that value mean? Complete
   this sentence: "If we randomly pick one customer who responded and
   one who did not, the model assigns a higher predicted probability
   to the responder \_\_\_\_% of the time."
3. The ROC curve plot shows the sensitivity/specificity tradeoff across
   all thresholds. At what point on the curve would you operate if your
   campaign budget only allows you to mail the top 10% of prospects?


AUC Value: The AUC value is approximately 0.473. This is below the 0.65 threshold, indicating that the model is not usable for targeting.
Interpretation of AUC: An AUC of 0.473 means that if we randomly pick one customer who responded and one who did not, the model assigns a higher predicted probability to the responder 47.3% of the time. This indicates that the model is performing worse than random guessing.
Operating Point on the ROC Curve: If your campaign budget only allows you to mail the top 10% of prospects, you would typically operate at the point on the ROC curve that corresponds to the highest true positive rate (sensitivity) while keeping the false positive rate low. This point is often found by determining the threshold that captures the top 10% of predicted probabilities. You would look for the threshold where the cumulative proportion of positives reaches 10% of the total population.

```{r ex9-2a, eval=FALSE}
# Your code here
auc_val
```

**Part B — Custom `metric_set()`**

Create a `metric_set()` containing `accuracy`, `roc_auc`, `sens`, `spec`,
and `f_meas`. Apply it to `test_preds`. If you had to report just **two**
of these five numbers to a VP of Marketing who is unfamiliar with
machine learning metrics, which would you choose and why?

```{r ex9-2b, eval=FALSE}
# Your code here
library(yardstick)

# Create a metric set containing accuracy, roc_auc, sens, spec, and f_meas
metric_set <- metric_set(accuracy, roc_auc, sens, spec, f_meas)

# Apply the metric set to test_preds
results <- metric_set(test_preds, truth = responded, estimate = .pred_class)

results
```

If I had to report just two of these five numbers to a VP of Marketing who is unfamiliar with machine learning metrics, I would choose:

Accuracy: This is a straightforward metric that indicates the overall performance of the model in terms of correct predictions. It is easy to understand and provides a general sense of how well the model is performing.

ROC AUC: This metric provides insight into the model's ability to distinguish between responders and non-responders. It is particularly useful for understanding the trade-offs between sensitivity and specificity across different thresholds.

## ✏️ Exercise 9.3 — Regression Metric Sensitivity to Outliers

Add one synthetic "whale" customer to `reg_preds` with `order_value_next = 5000` and `.pred = 200` (a badly missed prediction). Recompute `rmse` and `mae`. Which metric moved more in percentage terms? What does this tell you about choosing RMSE vs. MAE for marketing data with occasional huge orders?

```{r ex9-3, eval=FALSE}
# Your code here

# Add a synthetic 'whale' customer to reg_preds
ew_customer <- data.frame(order_value_next = 5000, .pred = 200)
reg_preds <- rbind(reg_preds, new_customer)

# Recompute RMSE and MAE
rmse_before <- sqrt(mean((reg_preds$order_value_next - reg_preds$.pred)^2))
mae_before <- mean(abs(reg_preds$order_value_next - reg_preds$.pred))

# Calculate RMSE and MAE after adding the whale customer
rmse_after <- sqrt(mean((reg_preds$order_value_next - reg_preds$.pred)^2))
mae_after <- mean(abs(reg_preds$order_value_next - reg_preds$.pred))

# Calculate percentage change
rmse_change <- ((rmse_after - rmse_before) / rmse_before) * 100
mae_change <- ((mae_after - mae_before) / mae_before) * 100

list(rmse_before = rmse_before, rmse_after = rmse_after, rmse_change = rmse_change,
     mae_before = mae_before, mae_after = mae_after, mae_change = mae_change)

```

## ✏️ Exercise 9.4 — Multiclass Averaging

Re-run the multiclass `loyalty_tier` example using `estimator = "micro"`
instead of the default macro averaging for `sens`. Also run it with
`estimator = "macro_weighted"`. Compare all three numbers.

1. Why might micro-averaging look more favorable when one class (Bronze)
   is much larger than the others?
2. Which estimator would you choose if you wanted to ensure that
   **Platinum** customers — the rarest and most valuable tier — are
   not underrepresented in the metric? Explain your reasoning.
3. In a marketing context, which averaging method is most likely to give
   a misleadingly optimistic picture of model performance? Why?


Why might micro-averaging look more favorable when one class (Bronze) is much larger than the others?
Micro-averaging calculates metrics globally by counting the total true positives, false negatives, and false positives across all classes. When one class (like Bronze) is much larger, its performance heavily influences the overall metrics. If the model performs well on the Bronze class, the micro-averaged metrics (like precision, recall, and F1-score) will appear more favorable, potentially masking poor performance on the smaller classes. This can lead to an overly optimistic view of the model's performance.
Which estimator would you choose if you wanted to ensure that Platinum customers — the rarest and most valuable tier — are not underrepresented in the metric? Explain your reasoning.
To ensure that Platinum customers are not underrepresented, you would choose a macro-averaging estimator. Macro-averaging calculates metrics for each class independently and then takes the average, treating all classes equally regardless of their size. This approach ensures that the performance on the Platinum class is given equal weight, preventing it from being overshadowed by the larger classes. This is crucial in a marketing context where the Platinum tier represents high-value customers.
In a marketing context, which averaging method is most likely to give a misleadingly optimistic picture of model performance? Why?
Micro-averaging is most likely to give a misleadingly optimistic picture of model performance in a marketing context. This is because it aggregates the performance across all classes, which can be heavily influenced by the performance on the majority class (e.g., Bronze). If the model performs well on the larger class but poorly on the smaller, more valuable classes (like Platinum), the micro-averaged metrics may suggest that the model is performing well overall, even though it fails to effectively target the high-value segments. This can lead to misguided marketing strategies and resource allocation.

# Appendix

Github Repository: https://github.com/brianvuong1/r-project-git

