---
title: "M05: Reflection on Ch09 (Evaluation)"
author: "Brian V"
date: today
format:

  html:
    toc: true
    toc-depth: 4
    toc-expand: 1
    toc-location: right-body
    toc-title: "Contents"
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    highlight-style: github
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
  freeze: true
---



# Exercises

## ✏️ Exercise 9.1 — Confusion Matrix and Threshold Tradeoff

**Part A — Reading the Confusion Matrix**

Run `conf_mat(test_preds, truth = responded, estimate = .pred_class)` and
use the output to answer:

1. How many customers actually responded ("yes") and were correctly
   predicted to respond? What is this cell called?
2. How many customers actually responded but were predicted "no"?
   What is this cell called, and what does it cost a marketer?
3. How many customers did not respond but were predicted "yes"?
   What is this cell called, and what does it cost a marketer?
4. In plain English, which type of error is more costly for a
   direct-mail campaign with a $5 mailing cost and $50 average order
   value — a false positive or a false negative? Explain your reasoning.


True Positives: 0 customers. This cell is called True Positives (TP).
False Negatives: 31 customers. This cell is called False Negatives (FN). The cost to a marketer for a false negative is the potential revenue lost from these customers who actually wanted to respond but were not targeted. If the average order value is $50, the cost would be (31 \times 50 = 1550).
False Positives: 0 customers. This cell is called False Positives (FP). The cost to a marketer for a false positive is the cost of mailing to customers who did not respond. If the mailing cost is $5, the cost would be (0 \times 5 = 0).
Costly Error: A false negative is more costly for a direct-mail campaign in this scenario. This is because the marketer misses out on potential sales from customers who would have responded positively, leading to a loss of revenue. In contrast, a false positive incurs only the cost of mailing without generating any revenue, which is less impactful compared to the lost sales from false negatives.

```{r ex9-1a, eval=FALSE}
# Your code here
conf_mat(test_preds, truth = responded, estimate = .pred_class)

```

**Part B — Threshold Tradeoff**

Using `test_preds`, compute `precision` and `sens` at the default 0.5
threshold. Now manually create a new `.pred_class` column using a threshold
of 0.3 instead (classify as "yes" if `.pred_yes > 0.3`). Recompute both
metrics. Which one increased, and why does lowering the threshold have
that effect?

Observations:
Both precision and sensitivity remain at 0 after lowering the threshold to 0.3. This indicates that even with the new threshold, there are still no true positives predicted.

Explanation:
Lowering the threshold generally increases sensitivity because it allows more instances to be classified as positive (in this case, "yes"). However, in this specific scenario, since there were no true positives to begin with, both metrics did not change.
In a typical case, lowering the threshold would lead to an increase in sensitivity (more true positives) but could decrease precision (more false positives). However, since we have no true positives in either case, both metrics remain at zero.

```{r ex9-1b, eval=FALSE}
# Calculate precision and sensitivity at the default threshold of 0.5
TP <- confusion_matrix$True_Positive
FN <- confusion_matrix$False_Negative
FP <- confusion_matrix$False_Positive

# Precision and Sensitivity calculations
precision_default <- ifelse((TP + FP) > 0, TP / (TP + FP), 0)
sensitivity_default <- ifelse((TP + FN) > 0, TP / (TP + FN), 0)

# Create a new .pred_class column using a threshold of 0.3
# Classify as 'yes' if .pred_yes > 0.3
test_preds <- test_preds %>% 
  mutate(.pred_class = ifelse(.pred_yes > 0.3, 'yes', 'no'))

# Recalculate confusion matrix components with the new threshold
new_confusion_matrix <- test_preds %>% 
  summarize(
    True_Positive = sum(responded == 'yes' & .pred_class == 'yes'),
    False_Negative = sum(responded == 'yes' & .pred_class == 'no'),
    False_Positive = sum(responded == 'no' & .pred_class == 'yes')
  )

new_confusion_matrix

precision_default
sensitivity_default

# Recalculate precision and sensitivity with the new threshold
TP_new <- new_confusion_matrix$True_Positive
FN_new <- new_confusion_matrix$False_Negative
FP_new <- new_confusion_matrix$False_Positive

# Precision and Sensitivity calculations for new threshold
precision_new <- ifelse((TP_new + FP_new) > 0, TP_new / (TP_new + FP_new), 0)
sensitivity_new <- ifelse((TP_new + FN_new) > 0, TP_new / (TP_new + FN_new), 0)

precision_new


```

## ✏️ Exercise 9.2 — ROC/AUC Interpretation and Custom `metric_set()`

**Part A — Interpreting AUC**

The notebook computed `roc_auc` for the direct-mail response model.

1. Look up the AUC value from the `roc-curve` chunk. Is it above or
   below 0.65 (the "usable for targeting" threshold from the callout)?
2. In plain English, what does an AUC of that value mean? Complete
   this sentence: "If we randomly pick one customer who responded and
   one who did not, the model assigns a higher predicted probability
   to the responder \_\_\_\_% of the time."
3. The ROC curve plot shows the sensitivity/specificity tradeoff across
   all thresholds. At what point on the curve would you operate if your
   campaign budget only allows you to mail the top 10% of prospects?


AUC Value: The AUC value is approximately 0.473. This is below the 0.65 threshold, indicating that the model is not usable for targeting.
Interpretation of AUC: An AUC of 0.473 means that if we randomly pick one customer who responded and one who did not, the model assigns a higher predicted probability to the responder 47.3% of the time. This indicates that the model is performing worse than random guessing.
Operating Point on the ROC Curve: If your campaign budget only allows you to mail the top 10% of prospects, you would typically operate at the point on the ROC curve that corresponds to the highest true positive rate (sensitivity) while keeping the false positive rate low. This point is often found by determining the threshold that captures the top 10% of predicted probabilities. You would look for the threshold where the cumulative proportion of positives reaches 10% of the total population.

```{r ex9-2a, eval=FALSE}
# Your code here
auc_val
```

**Part B — Custom `metric_set()`**

Create a `metric_set()` containing `accuracy`, `roc_auc`, `sens`, `spec`,
and `f_meas`. Apply it to `test_preds`. If you had to report just **two**
of these five numbers to a VP of Marketing who is unfamiliar with
machine learning metrics, which would you choose and why?

```{r ex9-2b, eval=FALSE}
# Your code here
library(yardstick)

# Create a metric set containing accuracy, roc_auc, sens, spec, and f_meas
metric_set <- metric_set(accuracy, roc_auc, sens, spec, f_meas)

# Apply the metric set to test_preds
results <- metric_set(test_preds, truth = responded, estimate = .pred_class)

results
```

If I had to report just two of these five numbers to a VP of Marketing who is unfamiliar with machine learning metrics, I would choose:

Accuracy: This is a straightforward metric that indicates the overall performance of the model in terms of correct predictions. It is easy to understand and provides a general sense of how well the model is performing.

ROC AUC: This metric provides insight into the model's ability to distinguish between responders and non-responders. It is particularly useful for understanding the trade-offs between sensitivity and specificity across different thresholds.

## ✏️ Exercise 9.3 — Regression Metric Sensitivity to Outliers

Add one synthetic "whale" customer to `reg_preds` with `order_value_next = 5000` and `.pred = 200` (a badly missed prediction). Recompute `rmse` and `mae`. Which metric moved more in percentage terms? What does this tell you about choosing RMSE vs. MAE for marketing data with occasional huge orders?

```{r ex9-3, eval=FALSE}
# Your code here

# Add a synthetic 'whale' customer to reg_preds
ew_customer <- data.frame(order_value_next = 5000, .pred = 200)
reg_preds <- rbind(reg_preds, new_customer)

# Recompute RMSE and MAE
rmse_before <- sqrt(mean((reg_preds$order_value_next - reg_preds$.pred)^2))
mae_before <- mean(abs(reg_preds$order_value_next - reg_preds$.pred))

# Calculate RMSE and MAE after adding the whale customer
rmse_after <- sqrt(mean((reg_preds$order_value_next - reg_preds$.pred)^2))
mae_after <- mean(abs(reg_preds$order_value_next - reg_preds$.pred))

# Calculate percentage change
rmse_change <- ((rmse_after - rmse_before) / rmse_before) * 100
mae_change <- ((mae_after - mae_before) / mae_before) * 100

list(rmse_before = rmse_before, rmse_after = rmse_after, rmse_change = rmse_change,
     mae_before = mae_before, mae_after = mae_after, mae_change = mae_change)

```

## ✏️ Exercise 9.4 — Multiclass Averaging

Re-run the multiclass `loyalty_tier` example using `estimator = "micro"`
instead of the default macro averaging for `sens`. Also run it with
`estimator = "macro_weighted"`. Compare all three numbers.

1. Why might micro-averaging look more favorable when one class (Bronze)
   is much larger than the others?
2. Which estimator would you choose if you wanted to ensure that
   **Platinum** customers — the rarest and most valuable tier — are
   not underrepresented in the metric? Explain your reasoning.
3. In a marketing context, which averaging method is most likely to give
   a misleadingly optimistic picture of model performance? Why?


Why might micro-averaging look more favorable when one class (Bronze) is much larger than the others?
Micro-averaging calculates metrics globally by counting the total true positives, false negatives, and false positives across all classes. When one class (like Bronze) is much larger, its performance heavily influences the overall metrics. If the model performs well on the Bronze class, the micro-averaged metrics (like precision, recall, and F1-score) will appear more favorable, potentially masking poor performance on the smaller classes. This can lead to an overly optimistic view of the model's performance.
Which estimator would you choose if you wanted to ensure that Platinum customers — the rarest and most valuable tier — are not underrepresented in the metric? Explain your reasoning.
To ensure that Platinum customers are not underrepresented, you would choose a macro-averaging estimator. Macro-averaging calculates metrics for each class independently and then takes the average, treating all classes equally regardless of their size. This approach ensures that the performance on the Platinum class is given equal weight, preventing it from being overshadowed by the larger classes. This is crucial in a marketing context where the Platinum tier represents high-value customers.
In a marketing context, which averaging method is most likely to give a misleadingly optimistic picture of model performance? Why?
Micro-averaging is most likely to give a misleadingly optimistic picture of model performance in a marketing context. This is because it aggregates the performance across all classes, which can be heavily influenced by the performance on the majority class (e.g., Bronze). If the model performs well on the larger class but poorly on the smaller, more valuable classes (like Platinum), the micro-averaged metrics may suggest that the model is performing well overall, even though it fails to effectively target the high-value segments. This can lead to misguided marketing strategies and resource allocation.

# Appendix

Github Repository: https://github.com/brianvuong1/r-project-git

Published Report: https://brianvuong1.github.io/r-project-git/W05-Reflection.html



